---
title: Novel Approaches for visualising Time to Event Health Data
author:
  - name: Victor Pacifique Rwandarwacu
  - name: Ranganath Shyamsundar 
    affil: 1
affiliation:
  - num: 1
    address: School of Mathematical and Statistical Sciences, University of Galway
logoleft_name: https&#58;//sport.universityofgalway.ie/assets/img/logo2.png

poster_height: "594mm"
poster_width: "841mm"
column_numbers: 4
column_margins: "0.5in"	
primary_colour:	"#a80050"	# Secondary colour use for poster design.
secondary_colour:	"#84003d"	# Main colour used for poster design.
accent_colour:	"#e6007e"	# A third colour option.
author_textcol: "white"
title_textsize: "70pt"
author_textsize: "50pt"
body_textsize: "26pt"
output: 
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false
knit: pagedown::chrome_print
---

```{css, echo=FALSE}
div.logo_left{
  width: 20%;
}
div.poster_title{
  width: 80%;
}
.section h4 {
    break-after: column;
}
div.footnotes {
    font-size: 20pt;
}
p.caption {
 margin-top: 0pt;
 margin-bottom: 8pt;
}
img {
 margin-top: 8pt;
 margin-bottom: 0pt;
}
```

<!-- Don't change anything above, except the title and author names, unless you know what you are doing. -->

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
# Load any additional libraries here
library(tidyverse)
library(plotly)
library(kableExtra)
library(posterdown)
library(htmltools)
# Load required libraries
library(survival)
library(ggplot2)
library(survminer)

# Load the veteran dataset
surv_brca_cli<- read.csv("surv_brca_cli.csv")
#knitr::opts_chunk$set(fig.topcaption = TRUE) #to keep all captions on the top
```

# Background and objectives

Time-to-event or survival Analysis is the analysis of data in the form of times from a well-defined time origin until the occurrence of some particular event or end point[^1]. Survival data are generally asymmetric and censored, which requires the use of specific approaches for analysis and visualisations, such as this survival function, Kaplan Meier(KM) estimator and plot.

[^1]: David Collett, Modelling survival data in medical research , Fourth Ed.

The survival function $S(t)$ is the probability that the survival time is greater than or equal to time $(t)$ which is the observed value of random variable $T$ with distribution function $F(t)$[^2].

[^2]: Peace, Karl E.. Design and Analysis of Clinical Trials with Time-to-Event Endpoints (Chapman & Hall/CRC Biostatistics Series) (p. 74). CRC Press. Kindle Edition.

$$
S(t)=\mathrm{P}(T \geqslant t)=1-F(t)   
$$

$$
F(t)=\mathrm{P}(T<t)=\int_0^t f(u) \mathrm{d} u
$$
The Kaplan Meier estimate of the survival function at $k$th interval is given by:

$$\hat{S}(t)=\prod_{j=1}^k\left(\frac{n_j-d_j}{n_j}\right)$$

For $t_{(k)} \leqslant t<t_{(k+1)}, k=1,2, \ldots, r$, with $\hat{S}(t)=1$ for $t<t_{(1)}$,where $t_{r+1}$ is taken to be $\infty$
$d_j$ denotes the number of deaths in this interval, $n_j$ is the number of individuals alive just before $t_{(j)}$ and $d_j$ deaths at $t_{(j)}$.

Survival Ratio, a robust approach for comparing survival distributions [^3], is defined by: $$R(t) = \frac{S_1(t)}{S_2(t)}$$

[^3]: J.Newell et.al, Survival ratio plots with permutation envelopes in survival data problems, <https://doi.org/10.1016/j.compbiomed.2005.03.005>

This project explores the use of novel informative visualisations of time-to-event data, specifically comparing survival curves of different covariates or treatments in a trial.



# Data Source and Dataset

The dataset is from the NIH National Cancer Institute , TCGA Program on a project called "Breast invasive carcinoma (BRCA)". It contains information about: demography, exposure , Family History(regarding cancer), Follow up, Molecular Test, other Clinical Attribute, pathology detail,and Treatment of female Breast cancer patients diagnosed and followed up for different outcomes.For demonstration, our analysis focuses on Survival outcomes by pathologic stages [^4]

[^4]: \<TCGA-BRCA , <https://portal.gdc.cancer.gov/projects/TCGA-BRCA>\>

# Proposed Approach

Table \@ref(tab:mytable) demonstrates the section of survival function and the change of the number of people at risk on each time interval.

```{r mytable, out.width='80%'}

surv_object <- Surv(time = surv_brca_cli$time_in_years, event = surv_brca_cli$vital_status)
fit1 <- survfit(surv_object ~ 1, data = surv_brca_cli)

mortality_summary <- data.frame(
    Time_Yrs = summary(fit1)$time,
    Survival_Prob = summary(fit1)$surv,
    n.risk= summary(fit1)$n.risk,
    Std.Error = summary(fit1)$std.err,
    Lower.95CI = summary(fit1)$lower,
    Upper.95CI = summary(fit1)$upper
  )


knitr::kable(slice(mortality_summary,100:105), caption = "<b style='font-size:22px;'>BRCA data survival</b>",align = 'c',"html") %>%kable_styling(font_size = 20, full_width = FALSE, position = "center") 

```

 
 
 

*Figure \@ref(fig:standard-plot), \@ref(fig:standard-plot1) and \@ref(fig:morefigs)*  highlight different approaches of visualising the estimated survival function.



```{r standard-plot, out.width='100%', fig.align='center', fig.cap='*KM plot_overall pathologic stages*   ', fig.height=5}


surv_object <- Surv(time = surv_brca_cli$time_in_years, event = surv_brca_cli$vital_status)
#fit the kaplan maier model for pathologic stage 
fit_path_stage <- survfit(surv_object ~ pathologic_stage , data = surv_brca_cli)

ggsurvplot(
  fit_path_stage,
  size = 1.2,                  # Line thickness
  font.x = c(16, "bold"),      # Font size for X-axis
  font.y = c(16, "bold"),      # Font size for Y-axis
  font.tickslab = c(10),       # Font size for tick labels
  legend = "right",             # Legend position
  conf.int = TRUE,                
  risk.table = TRUE,             
  pval = TRUE, 
  title="KM plot _Overall Pathologic stages",
 font.title = c(20, "bold"),
  xlab = "Time(years)",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal() + theme(
      legend.title = element_text(size = 10),  # Reduce legend title size
      legend.text = element_text(size = 8),   # Reduce legend text size
      legend.key.size = unit(0.3, "cm")       # Reduce legend key size
    ),      
  legend.title = "pathological stage",     
  risk.table.height = 0.4,        
  surv.median.line = "hv",
   width = 15,              # Width of the graph (in inches)
  height = 8               # Height of the graph (in inches)
  
)


```

```{r standard-plot1, out.width='100%', fig.align='center', fig.cap='*KM plot Of Pathologic stage II and III* ', fig.height=5}
path_stage2_3_df<- surv_brca_cli|> select(time,vital_status,pathologic_stage)|> filter(pathologic_stage == "Stage II" | pathologic_stage == "Stage III")

fit23 <- survfit(Surv(time,vital_status)~pathologic_stage, data = path_stage2_3_df)


ggsurvplot(
  fit23,
  conf.int = TRUE,                
  risk.table = TRUE,
  title = "KM plot of pathologic stage II and III",
  font.title = c(20, "bold"),
  xlab = "Time in days",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),
  pval = TRUE,
  risk.table.height = 0.3,
  risk.table.length = 0.5
  
)  
  

```

```{r morefigs, out.width='100%', echo=FALSE, fig.cap='*Survival Ratio plot for Path. stage II/ III with 95% C.I*', fig.height=5}

plot_data<-read.csv("plot_data.csv")

ggplot(plot_data, aes(x = time)) +
  # Survival ratio line
  geom_line(aes(y = ratio, color = "Survival Ratio"), size = 1.2) +
  
  # Confidence interval ribbon
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  
  # Reference line
  geom_hline(aes(yintercept = 1, linetype = "No Difference Line"), color = "red", size = 1 ) +
  
  
  # Custom scales for the legend
  scale_color_manual(
    name = "Line",
    values = c("Survival Ratio" = "green")
  ) +
  scale_fill_manual(
    name = "Confidence Bands",
    values = c(
      "95% Confidence Interval" = "blue"
    )
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("No Difference Line" = "dashed")
  ) +
  
  # Labels and theme
  labs(
    title = "Survival Ratio plot for Path. stage II/ III with Confidence intervals",
    x = "Time (Days)",
    y = "Survival Ratio (Pathologic Stage II / Stage III)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  # Position of the legend
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )+ 
  annotate("text", x = max(plot_data$Time) * 0.3, y = 2, label = "Higher survival for Pathologic stage 2", hjust = 0) +
  annotate("text", x = max(plot_data$Time) * 0.3, y = 0.5, label = "Higher survival for pathologic stage 3", hjust = 0) 




```


### 

# Next  Steps

-Visualize survival differences between independent groups, incorporating confidence intervals to assess variability and significance.

-Generate survival ratio plots for paired data, using permutation envelopes as reference bands to evaluate deviations and provide robust comparisons.

-Compare survival distributions across more than two groups utilizing non-parametric statistical methods to identify significant differences.

# GitHub

The code and dataset for this project can be found at  GitHub repository through this link : <https://github.com/rwandarwacu1/Msc_thesis_survival>

# References
