---
title: "Plots for survey and Manuscript"
author: "Victor Rwandarwacu"
date: "2025-01-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## INTRODUCTION

This document includes Kaplan Meier and survival ratio plots to be used for survey among clinician.

## Rationale for the project

Standard visualisation of survival data enable understand of time to event and survival probabilty at certain period. however , there is limitation when it comes to comparison of different groups.

\~literature review

This project aims at generating comparative plots of survival probabilities for both paired and unpaired data.

## DATA SETS

TCGA BRCA

```{r }
# Required libraries
library(survival)
library(survminer)
library(ggplot2)
library(tidyverse)



surv_brca_cli<- read.csv("surv_brca_cli.csv")
surv_object <- Surv(time = surv_brca_cli$time_in_years, event = surv_brca_cli$vital_status)
```

## KM Plot including all covariates

```{r}
#fit the kaplan maier model for pathologic stage 
fit_path_stage <- survfit(surv_object ~ pathologic_stage , data = surv_brca_cli)

ggsurvplot(
  fit_path_stage,
  size = 1.5,                  # Line thickness
  font.x = c(16, "bold"),      # Font size for X-axis
  font.y = c(16, "bold"),      # Font size for Y-axis
  font.tickslab = c(10),       # Font size for tick labels
  legend = "right",            
  conf.int = TRUE,                
  risk.table = TRUE,             
  pval = TRUE,                    
  xlab = "Time(years)",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),      
  legend.title = "pathological stage",     
  risk.table.height = 0.5,        
  surv.median.line = "hv" 
  
)


```

#################################################################################################### 

## **COMPARISON OF PATHOLOGIC STAGE 2 AND 3**

#################################################################################################### 

## Kaplan Meier plot for stage 2 and stage 3

```{r}
path_stage2_3_df<- surv_brca_cli|> select(time_in_years,vital_status,pathologic_stage)|> filter(pathologic_stage == "Stage II" | pathologic_stage == "Stage III")

fit23 <- survfit(Surv(time_in_years,vital_status)~pathologic_stage, data = path_stage2_3_df)


ggsurvplot(
  fit23,
  conf.int = TRUE,                
  risk.table = TRUE,              
  xlab = "Time in years",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),
  pval = FALSE,
  title= "Kaplan Meier plot for pathologic stage II and III"
  
)  
  


```

```{r}

#KM fit of the data
km_fit_path_stageII <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_stage == "Stage II"))

km_fit_path_stageIII <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_stage == "Stage III"))

####extract survival probabilities 

# Manually define time points (we mentained the exact time in the data set in days )
manual_times <- seq(0, max(surv_brca_cli$time_in_years, na.rm = TRUE), by = 0.08333)
# Extract survival probabilities for Pathologic stage 2
surv_pathII <- summary(km_fit_path_stageII, times = manual_times, extend = TRUE)$surv

# Extract survival probabilities for pathologic stage 3
surv_pathIII <- summary(km_fit_path_stageIII, times = manual_times, extend = TRUE)$surv
# Survival ratios and confidence intervals
survival_ratio <- surv_pathII / surv_pathIII

```

## Confidence interval estimation

This confidence interval is calculated using a delta method variance is calculated from the standard error of survival probabilities

```{r}

# Confidence intervals using the Delta method


# Extract Variances  for Pathologic stage 2 and 3
var_pathII <- (summary(km_fit_path_stageII, times = manual_times, extend = TRUE)$std.err)^2
var_pathIII <- (summary(km_fit_path_stageIII, times = manual_times, extend = TRUE)$std.err)^2



# Log-transformed survival ratio
log_ratio <- log(survival_ratio)

# Variance of the log survival ratio
log_ratio_var <- var_pathIII / (surv_pathIII^2) + var_pathII / (surv_pathII^2)

# Confidence intervals for log-ratio
z_value <- qnorm(0.975)  
lower_log <- log_ratio - z_value * sqrt(log_ratio_var)
upper_log <- log_ratio + z_value * sqrt(log_ratio_var)

# Back-transform to get confidence intervals for the ratio
lower_ci <- exp(lower_log)
upper_ci <- exp(upper_log)

# Add CI to the plot
plot_data <- data.frame(
  time = manual_times,
  ratio = survival_ratio,
  lower_ci = lower_ci,
  upper_ci = upper_ci
  
)

#write.csv(plot_data,"plot_data.csv",row.names = FALSE)







```

## Plot data and ratio plot with confidence interval

```{r}

ggplot(plot_data, aes(x = time)) +
  # Survival ratio line
  geom_line(aes(y = ratio, color = "Survival Ratio"), size = 1.2) +
  
  # Confidence interval ribbon
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  
  # Reference line
  geom_hline(aes(yintercept = 1, linetype = "No Difference Line"), color = "red", size = 1 ) +
  
  
  # Custom scales for the legend
  scale_color_manual(
    name = "Line",
    values = c("Survival Ratio" = "green")
  ) +
  scale_fill_manual(
    name = "Confidence Bands",
    values = c(
      "95% Confidence Interval" = "blue"
    )
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("No Difference Line" = "dashed")
  ) +
  
  # Labels and theme
  labs(
    title = "Survival Ratio plot for Path. stage II/ III ",
    x = "Time (years)",
    y = "Survival Ratio (Pathologic Stage II / Stage III)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  # Position of the legend
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )+ 
  annotate("text", x = max(plot_data$Time) * 0.3, y = 2, label = "Higher survival for Pathologic stage 2", hjust = 0) +
  annotate("text", x = max(plot_data$Time) * 0.3, y = 0.5, label = "Higher survival for pathologic stage 3", hjust = 0)  





```

## SURVIVAL DIFFERENCE PLOT STAGE 2 AND 3

```{r}

# Compute survival difference
survival_diff_stage2_3 <- surv_pathII - surv_pathIII

# Compute standard error for difference
diff_var_stage2_3 <- var_pathII + var_pathIV
z_value <- qnorm(0.975)
lower_ci_stage2_3dif <- survival_diff_stage2_3 - z_value * sqrt(diff_var_stage2_3)
upper_ci_stage2_3dif<- survival_diff_stage2_3 + z_value * sqrt(diff_var_stage2_3)

# Prepare data for plotting
plot_data_diff23 <- data.frame(
  time = manual_times,
  difference = survival_diff_stage2_3,
  lower_ci = lower_ci_stage2_3dif,
  upper_ci = upper_ci_stage2_3dif
)

# Generate survival difference plot
ggplot(plot_data_diff23, aes(x = time)) +
  geom_line(aes(y = difference, color = "Survival Difference"), size = 1.2) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  geom_hline(aes(yintercept = 0, linetype = "No Difference Line"), color = "red", size = 1) +
  scale_color_manual(name = "Line", values = c("Survival Difference" = "green")) +
  scale_fill_manual(name = "Confidence Bands", values = c("95% Confidence Interval" = "blue")) +
  scale_linetype_manual(name = "Reference", values = c("No Difference Line" = "dashed")) +
  labs(
    title = "Survival Difference Plot for Pathologic Stage II vs. III",
    x = "Time (Years)",
    y = "Survival Difference (Pathologic Stage II - Stage III)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )







```

#################################################################################################### 

## **COMPARISON OF PATHOLOGIC STAGE 2 AND 4**

#################################################################################################### 

## Kaplan Meier plot for stage 2 and stage 4

```{r}
path_stage2_4_df<- surv_brca_cli|> select(time_in_years,vital_status,pathologic_stage)|> filter(pathologic_stage == "Stage II" | pathologic_stage == "Stage IV")

fit24 <- survfit(Surv(time_in_years,vital_status)~pathologic_stage, data = path_stage2_4_df)


ggsurvplot(
  fit24,
  conf.int = TRUE,                
  risk.table = TRUE,              
  xlab = "Time in years",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),
  pval = FALSE,
  title= "Kaplan Meier plot for pathologic stage II and IV"
)  
```

```{r}


```

## RATIO PLOT STAGE 2 OVER 4

```{r}
data_path4 = subset(surv_brca_cli, pathologic_stage == "Stage IV")
```

```{r}



#KM fit of the data
km_fit_path_stageII <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_stage == "Stage II"))

km_fit_path_stageIV <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_stage == "Stage IV"))

####extract survival probabilities 

# Manually define time points (we mentained the exact time in the data set in days )
manual_times24 <- seq(0, max(data_path4$time_in_years, na.rm = TRUE), by = 0.083)
# Extract survival probabilities for Pathologic stage 2
surv_pathII <- summary(km_fit_path_stageII, times = manual_times24, extend = TRUE)$surv

# Extract survival probabilities for pathologic stage 3
surv_pathIV <- summary(km_fit_path_stageIV, times = manual_times24, extend = TRUE)$surv
# Survival ratios and confidence intervals
survival_ratio_stage2_4 <- surv_pathII / surv_pathIV

```

## Confidence interval estimation 2/4

```{r}

# Confidence intervals using the Delta method


# Extract Variances  for Pathologic stage 2 and 3
var_pathII <- (summary(km_fit_path_stageII, times = manual_times24, extend = TRUE)$std.err)^2
var_pathIV <- (summary(km_fit_path_stageIV, times = manual_times24, extend = TRUE)$std.err)^2



# Log-transformed survival ratio
log_ratio_stage2_4 <- log(survival_ratio_stage2_4)

# Variance of the log survival ratio
log_ratio_var_stage2_4 <- var_pathIV / (surv_pathIV^2) + var_pathII / (surv_pathII^2)

# Confidence intervals for log-ratio
z_value <- qnorm(0.975)  
lower_log_stage2_4 <- log_ratio_stage2_4 - z_value * sqrt(log_ratio_var_stage2_4)
upper_log_stage2_4 <- log_ratio_stage2_4 + z_value * sqrt(log_ratio_var_stage2_4)

# Back-transform to get confidence intervals for the ratio
lower_ci_stage2_4 <- exp(lower_log_stage2_4)
upper_ci_stage2_4 <- exp(upper_log_stage2_4)

# Add CI to the plot
plot_data_stage2_4 <- data.frame(
  time = manual_times24,
  ratio = survival_ratio_stage2_4,
  lower_ci = lower_ci_stage2_4,
  upper_ci = upper_ci_stage2_4
  
)




```

## Plot data and ratio plot with confidence interval

```{r}

ggplot(plot_data_stage2_4, aes(x = time)) +
  # Survival ratio line
  geom_line(aes(y = ratio, color = "Survival Ratio"), size = 1.2) +
  
  # Confidence interval ribbon
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  
  # Reference line
  geom_hline(aes(yintercept = 1, linetype = "No Difference Line"), color = "red", size = 1 ) +
  
  
  # Custom scales for the legend
  scale_color_manual(
    name = "Line",
    values = c("Survival Ratio" = "green")
  ) +
  scale_fill_manual(
    name = "Confidence Bands",
    values = c(
      "95% Confidence Interval" = "blue"
    )
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("No Difference Line" = "dashed")
  ) +
  
  # Labels and theme
  labs(
    title = "Survival Ratio plot for Path. stage II/ IV ",
    x = "Time (Years)",
    y = "Survival Ratio (Pathologic Stage II / Stage IV)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  # Position of the legend
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )+ 
  annotate("text", x = max(plot_data_stage2_4$Time) * 0.3, y = 14, label = "Higher survival for Pathologic stage 2", hjust = 0) +
  annotate("text", x = max(plot_data_stage2_4$Time) * 0.3, y = 0.5, label = "Higher survival for pathologic stage 4", hjust = 0)  





```

##comaprison survival diff stage 2 over 4

```{r}
# Compute survival difference
survival_diff_stage2_4 <- surv_pathII - surv_pathIV

# Compute standard error for difference
diff_var_stage2_4 <- var_pathII + var_pathIV
z_value <- qnorm(0.975)
lower_ci_stage2_4dif <- survival_diff_stage2_4 - z_value * sqrt(diff_var_stage2_4)
upper_ci_stage2_4dif<- survival_diff_stage2_4 + z_value * sqrt(diff_var_stage2_4)

# Prepare data for plotting
plot_data_diff24 <- data.frame(
  time = manual_times24,
  difference = survival_diff_stage2_4,
  lower_ci = lower_ci_stage2_4dif,
  upper_ci = upper_ci_stage2_4dif
)

# Generate survival difference plot
ggplot(plot_data_diff, aes(x = time)) +
  geom_line(aes(y = difference, color = "Survival Difference"), size = 1.2) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  geom_hline(aes(yintercept = 0, linetype = "No Difference Line"), color = "red", size = 1) +
  scale_color_manual(name = "Line", values = c("Survival Difference" = "green")) +
  scale_fill_manual(name = "Confidence Bands", values = c("95% Confidence Interval" = "blue")) +
  scale_linetype_manual(name = "Reference", values = c("No Difference Line" = "dashed")) +
  labs(
    title = "Survival Difference Plot for Pathologic Stage II vs. IV",
    x = "Time (Years)",
    y = "Survival Difference (Pathologic Stage II - Stage IV)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )


```

#################################################################################################### 

## **COMPARISON OF PATHOLOGIC STAGE 1 AND 2**

#################################################################################################### 

## Kaplan Meier plot for stage 1 and stage 2

```{r}
path_stage1_2_df<- surv_brca_cli|> select(time_in_years,vital_status,pathologic_stage)|> filter(pathologic_stage == "Stage I" | pathologic_stage == "Stage II")

fit12 <- survfit(Surv(time_in_years,vital_status)~pathologic_stage, data = path_stage1_2_df)


ggsurvplot(
  fit12,
  conf.int = TRUE,                
  risk.table = TRUE,              
  xlab = "Time in years",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),
  pval = FALSE,
  title= "Kaplan Meier plot for pathologic stage I and stage II"
)  
```

```{r}


```

## RATIO PLOT STAGE 1 OVER 2

```{r}

#KM fit of the data
km_fit_path_stageII <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_stage == "Stage II"))

km_fit_path_stageI <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_stage == "Stage I"))

####extract survival probabilities 

# Manually define time points (we mentained the exact time in the data set in days )
manual_times <- seq(0, max(surv_brca_cli$time_in_years, na.rm = TRUE), by = 0.083)
# Extract survival probabilities for Pathologic stage 2
surv_pathII <- summary(km_fit_path_stageII, times = manual_times, extend = TRUE)$surv

# Extract survival probabilities for pathologic stage 3
surv_pathI <- summary(km_fit_path_stageI, times = manual_times, extend = TRUE)$surv
# Survival ratios and confidence intervals
survival_ratio_stage1_2 <- surv_pathI / surv_pathII

```

## Confidence interval estimation 1/2

```{r}

# Confidence intervals using the Delta method


# Extract Variances  for Pathologic stage 2 and 3
var_pathII <- (summary(km_fit_path_stageII, times = manual_times, extend = TRUE)$std.err)^2
var_pathI <- (summary(km_fit_path_stageI, times = manual_times, extend = TRUE)$std.err)^2



# Log-transformed survival ratio
log_ratio_stage1_2 <- log(survival_ratio_stage1_2)

# Variance of the log survival ratio
log_ratio_var_stage1_2 <- var_pathI / (surv_pathI^2) + var_pathII / (surv_pathII^2)

# Confidence intervals for log-ratio
z_value <- qnorm(0.975)  
lower_log_stage1_2 <- log_ratio_stage1_2 - z_value * sqrt(log_ratio_var_stage1_2)
upper_log_stage1_2 <- log_ratio_stage1_2 + z_value * sqrt(log_ratio_var_stage1_2)

# Back-transformation to get confidence intervals for the ratio
lower_ci_stage1_2 <- exp(lower_log_stage1_2)
upper_ci_stage1_2 <- exp(upper_log_stage1_2)

# Add CI to the plot
plot_data_stage1_2 <- data.frame(
  time = manual_times,
  ratio = survival_ratio_stage1_2,
  lower_ci = lower_ci_stage1_2,
  upper_ci = upper_ci_stage1_2
  
)









```

## Plot data and ratio plot with confidence interval

```{r}

ggplot(plot_data_stage1_2, aes(x = time)) +
  
  geom_line(aes(y = ratio, color = "Survival Ratio"), size = 1.2) +
  
 
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  
  # Reference line
  geom_hline(aes(yintercept = 1, linetype = "No Difference Line"), color = "red", size = 1 ) +
  
  
  # Custom scales for the legend
  scale_color_manual(
    name = "Line",
    values = c("Survival Ratio" = "green")
  ) +
  scale_fill_manual(
    name = "Confidence Bands",
    values = c(
      "95% Confidence Interval" = "blue"
    )
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("No Difference Line" = "dashed")
  ) +
  
  # Labels and theme
  labs(
    title = "Survival Ratio plot for Path. stage I/ II ",
    x = "Time (Years)",
    y = "Survival Ratio (Pathologic Stage I / Stage II)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  # Position of the legend
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )+ 
  annotate("text", x = max(plot_data_stage1_2$Time) * 0.3, y = 2, label = "Higher survival for Pathologic stage 1", hjust = 0) +
  annotate("text", x = max(plot_data_stage1_2$Time) * 0.3, y = 0.5, label = "Higher survival for pathologic stage 2", hjust = 0)  





```

##### survival difference plot stage 1 and 2

```{r}

# Compute survival difference
survival_diff_stage1_2 <- surv_pathI - surv_pathII

# Compute standard error for difference
diff_var_stage1_2 <- var_pathI + var_pathII
z_value <- qnorm(0.975)
lower_ci_stage1_2dif <- survival_diff_stage1_2 - z_value * sqrt(diff_var_stage1_2)
upper_ci_stage1_2dif<- survival_diff_stage1_2 + z_value * sqrt(diff_var_stage1_2)

# Prepare data for plotting
plot_data_diff12 <- data.frame(
  time = manual_times,
  difference = survival_diff_stage1_2,
  lower_ci = lower_ci_stage1_2dif,
  upper_ci = upper_ci_stage1_2dif
)

# Generate survival difference plot
ggplot(plot_data_diff12, aes(x = time)) +
  geom_line(aes(y = difference, color = "Survival Difference"), size = 1.2) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  geom_hline(aes(yintercept = 0, linetype = "No Difference Line"), color = "red", size = 1) +
  scale_color_manual(name = "Line", values = c("Survival Difference" = "green")) +
  scale_fill_manual(name = "Confidence Bands", values = c("95% Confidence Interval" = "blue")) +
  scale_linetype_manual(name = "Reference", values = c("No Difference Line" = "dashed")) +
  labs(
    title = "Survival Difference Plot for Pathologic Stage I vs. II",
    x = "Time (Years)",
    y = "Survival Difference (Pathologic Stage I - Stage II)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )


```

#################################################################################################### 

## **COMPARISON OF SURVIVAL AMONG RACE** 

#################################################################################################### 

## Kaplan Meier plot for stage 2 and stage 4

```{r}
#fit the kaplan maier model 
race_data= surv_brca_cli %>% mutate(race= as.factor(race)) %>% filter(race != "Other")
surv_object_race= Surv(time = race_data$time_in_years, event = race_data$vital_status) 
fit_race <- survfit(surv_object_race ~ race , data = race_data)
#plot the survival curve 

ggsurvplot(
  fit_race,
  size = 1.5,                  
  font.x = c(12, "bold"),     
  font.y = c(12, "bold"),     
  font.tickslab = c(10),       
  legend = "right",             
  conf.int = TRUE,                
  risk.table = TRUE,             
  pval = FALSE,                    
  xlab = "Time(years)",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),      
  legend.title = "race",     
  risk.table.height = 0.3,        
  surv.median.line = "hv" ,
  title = "Kaplan Meier  plot White vs Black race" 
  
)
```

```{r}


```

## RATIO PLOT Black ror african american versus white

```{r}

#KM fit of the data
km_fit_race_white <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, race == "White"))

km_fit_race_black <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, race == "Black or African American"))

####extract survival probabilities 

# Manually define time points (we mentained the exact time in the data set in days )
manual_times <- seq(0, max(surv_brca_cli$time_in_years, na.rm = TRUE), by = 0.083)
# Extract survival probabilities for Pathologic stage 2
surv_white <- summary(km_fit_race_white, times = manual_times, extend = TRUE)$surv

# Extract survival probabilities for pathologic stage 3
surv_black <- summary(km_fit_race_black, times = manual_times, extend = TRUE)$surv
# Survival ratios and confidence intervals
survival_ratio_race <- surv_white / surv_black

```

## Confidence interval estimation 1/2

```{r}




# Extract Variances  for Pathologic stage 2 and 3
var_white <- (summary(km_fit_race_white, times = manual_times, extend = TRUE)$std.err)^2
var_black <- (summary(km_fit_race_black, times = manual_times, extend = TRUE)$std.err)^2



# Log-transformed survival ratio
log_ratio_race <- log(survival_ratio_race)

# Variance of the log survival ratio
log_ratio_var_race <- var_white / (surv_white^2) + var_black / (surv_white^2)

# Confidence intervals for log-ratio
z_value <- qnorm(0.975)  
lower_log_race <- log_ratio_race - z_value * sqrt(log_ratio_var_race)
upper_log_race <- log_ratio_race + z_value * sqrt(log_ratio_var_race)

# Back-transform to get confidence intervals for the ratio
lower_ci_race <- exp(lower_log_race)
upper_ci_race <- exp(upper_log_race)

# Add CI to the plot
plot_data_race <- data.frame(
  time = manual_times,
  ratio = survival_ratio_race,
  lower_ci = lower_ci_race,
  upper_ci = upper_ci_race
  
)



```

## Plot data and ratio plot with confidence interval

```{r}

ggplot(plot_data_race, aes(x = time)) +
  # Survival ratio line
  geom_line(aes(y = ratio, color = "Survival Ratio"), size = 1.2) +
  
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  
  geom_hline(aes(yintercept = 1, linetype = "No Difference Line"), color = "red", size = 1 ) +
  
  scale_color_manual(
    name = "Line",
    values = c("Survival Ratio" = "green")
  ) +
  scale_fill_manual(
    name = "Confidence Bands",
    values = c(
      "95% Confidence Interval" = "blue"
    )
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("No Difference Line" = "dashed")
  ) +
  
  
  labs(
    title = "Survival Ratio plot for White/Black ",
    x = "Time (Years)",
    y = "Survival Ratio (Pathologic Stage White/Black)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  # Position of the legend
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )+ 
  annotate("text", x = max(plot_data_stage2_4$Time) * 0.3, y = 2, label = "Higher survival for white ", hjust = 0) +
  annotate("text", x = max(plot_data_stage2_4$Time) * 0.3, y = 0.5, label = "Higher survival for Black or African American", hjust = 0)  





```

##COMPARISON OF SURVIVAL DIFFERENCE ON RACE

```{r}

# Compute survival difference
survival_diff_race <- surv_white - surv_black

# Compute standard error for difference
diff_var_race <- var_white + var_black
z_value <- qnorm(0.975)
lower_ci_race_dif <- survival_diff_race - z_value * sqrt(diff_var_race)
upper_ci_race_dif<- survival_diff_race + z_value * sqrt(diff_var_race)

# Prepare data for plotting
plot_data_diff_race <- data.frame(
  time = manual_times,
  difference = survival_diff_race,
  lower_ci = lower_ci_race_dif,
  upper_ci = upper_ci_race_dif
)

# Generate survival difference plot
ggplot(plot_data_diff_race, aes(x = time)) +
  geom_line(aes(y = difference, color = "Survival Difference"), size = 1.2) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  geom_hline(aes(yintercept = 0, linetype = "No Difference Line"), color = "red", size = 1) +
  scale_color_manual(name = "Line", values = c("Survival Difference" = "green")) +
  scale_fill_manual(name = "Confidence Bands", values = c("95% Confidence Interval" = "blue")) +
  scale_linetype_manual(name = "Reference", values = c("No Difference Line" = "dashed")) +
  labs(
    title = "Survival Difference Plot for White vs. Black Race",
    x = "Time (Years)",
    y = "Survival Difference (White - Black Race)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )





```

##PATHOLOGICAL LYMPH NODE INVOLVEMENT

```{r}
#fit the kaplan meier model 

surv_brca_cli <- surv_brca_cli %>%
  mutate(pathologic_n = as.factor(substr(pathologic_n, 1, 2)))

summary(surv_brca_cli$pathologic_n)

lymph_nodes = surv_brca_cli %>% filter(pathologic_n =="N0"|pathologic_n=="N3")

fit_path_ln <- survfit(Surv(time = time_in_years,event = vital_status) ~ pathologic_n , data = lymph_nodes)
#plot the survival curve 

ggsurvplot(
  fit_path_ln,
  size = 1.5,                  
  font.x = c(12, "bold"),     
  font.y = c(12, "bold"),     
  font.tickslab = c(10),       
  legend = "right",             
  conf.int = TRUE,                
  risk.table = TRUE,             
    pval = FALSE,                  
  xlab = "Time(years)",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),      
  legend.title = "Path.Lymph nodes",     
  risk.table.height = 0.3,        
  surv.median.line = "hv" ,
  title="Pathologic stage Lymph-node involvement "
  
)
```

```{r}

#KM fit of the data
km_fit_n0 <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_n == "N0"))

km_fit_n1 <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_n == "N1"))

manual_times <- seq(0, max(surv_brca_cli$time_in_years, na.rm = TRUE), by = 0.083)

surv_n0 <- summary(km_fit_n0, times = manual_times, extend = TRUE)$surv

surv_n1 <- summary(km_fit_n1, times = manual_times, extend = TRUE)$surv

survival_ratio_n0n1 <- surv_n0 / surv_n1


```

```{r}

var_n1 <- (summary(km_fit_n1, times = manual_times, extend = TRUE)$std.err)^2
var_n0 <- (summary(km_fit_n0, times = manual_times, extend = TRUE)$std.err)^2



# Log-transformed survival ratio
log_ratio_n0n1 <- log(survival_ratio_n0n1)

# Variance of the log survival ratio
log_ratio_var_n0n1 <- var_n1 / (surv_n1^2) + var_n0 / (surv_n0^2)

# Confidence intervals for log-ratio
z_value <- qnorm(0.975)  
lower_log_n0n1 <- log_ratio_n0n1 - z_value * sqrt(log_ratio_var_n0n1)
upper_log_n0n1 <- log_ratio_n0n1 + z_value * sqrt(log_ratio_var_n0n1)

# Back-transform to get confidence intervals for the ratio
lower_ci_n0n1 <- exp(lower_log_n0n1)
upper_ci_n0n1 <- exp(upper_log_n0n1)

# Add CI to the plot
plot_data_n0n1 <- data.frame(
  time = manual_times,
  ratio = survival_ratio_n0n1,
  lower_ci = lower_ci_n0n1,
  upper_ci = upper_ci_n0n1
  
)



```

```{r}

ggplot(plot_data_n0n1, aes(x = time)) +
  # Survival ratio line
  geom_line(aes(y = ratio, color = "Survival Ratio"), size = 1.2) +
  
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  
  geom_hline(aes(yintercept = 1, linetype = "No Difference Line"), color = "red", size = 1 ) +
  
  scale_color_manual(
    name = "Line",
    values = c("Survival Ratio" = "green")
  ) +
  scale_fill_manual(
    name = "Confidence Bands",
    values = c(
      "95% Confidence Interval" = "blue"
    )
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("No Difference Line" = "dashed")
  ) +
  
  
  labs(
    title = "Survival Ratio plot for N0/N1 ",
    x = "Time (Years)",
    y = "Survival Ratio (Pathologic N0/N1)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  # Position of the legend
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )+ 
  annotate("text", x = max(plot_data_stage2_4$Time) * 0.3, y = 2, label = "Higher survival N0 ", hjust = 0) +
  annotate("text", x = max(plot_data_stage2_4$Time) * 0.3, y = 0.5, label = "Higher survival for N1", hjust = 0)  





```

#SURVIVAL DIFFERENCE

```{r}
# Compute survival difference
survival_diff_n0n1 <- surv_n0 - surv_n1

# Compute standard error for difference
diff_var_n0n1 <- var_n0 + var_n1
z_value <- qnorm(0.975)
lower_ci_n0n1_dif <- survival_diff_n0n1 - z_value * sqrt(diff_var_n0n1)
upper_ci_n0n1_dif<- survival_diff_n0n1 + z_value * sqrt(diff_var_n0n1)

# Prepare data for plotting
plot_data_diff_n0n1 <- data.frame(
  time = manual_times,
  difference = survival_diff_n0n1,
  lower_ci = lower_ci_n0n1_dif,
  upper_ci = upper_ci_n0n1_dif
)

# Generate survival difference plot
ggplot(plot_data_diff_n0n1, aes(x = time)) +
  geom_line(aes(y = difference, color = "Survival Difference"), size = 1.2) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  geom_hline(aes(yintercept = 0, linetype = "No Difference Line"), color = "red", size = 1) +
  scale_color_manual(name = "Line", values = c("Survival Difference" = "green")) +
  scale_fill_manual(name = "Confidence Bands", values = c("95% Confidence Interval" = "blue")) +
  scale_linetype_manual(name = "Reference", values = c("No Difference Line" = "dashed")) +
  labs(
    title = "Survival Difference Plot TNM N0 VS N1 ",
    x = "Time (Years)",
    y = "Survival Difference (pathologic LN N0/N1)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )



```

##PATHOLOGICAL LYMPH NODE INVOLVEMENT NO/N3

```{r}
#fit the kaplan meier model 

surv_brca_cli <- surv_brca_cli %>%
  mutate(pathologic_n = as.factor(substr(pathologic_n, 1, 2)))

summary(surv_brca_cli$pathologic_n)

lymph_nodes03 = surv_brca_cli %>% filter(pathologic_n =="N0"|pathologic_n=="N2")

fit_path_ln <- survfit(Surv(time = time_in_years,event = vital_status) ~ pathologic_n , data = lymph_nodes03)
#plot the survival curve 

ggsurvplot(
  fit_path_ln,
  size = 1.5,                  
  font.x = c(12, "bold"),     
  font.y = c(12, "bold"),     
  font.tickslab = c(10),       
  legend = "right",             
  conf.int = TRUE,                
  risk.table = TRUE,             
    pval = FALSE,                  
  xlab = "Time(years)",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),      
  legend.title = "Path.Lymph nodes",     
  risk.table.height = 0.3,        
  surv.median.line = "hv" ,
  title="Pathologic stage Lymph-node involvement "
  
)
```

```{r}

#KM fit of the data
km_fit_n0 <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_n == "N0"))

km_fit_n1 <- survfit(Surv(time= time_in_years, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_n == "N2"))

manual_times <- seq(0, max(14, na.rm = TRUE), by = 0.083)

surv_n0 <- summary(km_fit_n0, times = manual_times, extend = TRUE)$surv

surv_n1 <- summary(km_fit_n1, times = manual_times, extend = TRUE)$surv

survival_ratio_n0n1 <- surv_n0 / surv_n1


```

```{r}

var_n1 <- (summary(km_fit_n1, times = manual_times, extend = TRUE)$std.err)^2
var_n0 <- (summary(km_fit_n0, times = manual_times, extend = TRUE)$std.err)^2



# Log-transformed survival ratio
log_ratio_n0n1 <- log(survival_ratio_n0n1)

# Variance of the log survival ratio
log_ratio_var_n0n1 <- var_n1 / (surv_n1^2) + var_n0 / (surv_n0^2)

# Confidence intervals for log-ratio
z_value <- qnorm(0.975)  
lower_log_n0n1 <- log_ratio_n0n1 - z_value * sqrt(log_ratio_var_n0n1)
upper_log_n0n1 <- log_ratio_n0n1 + z_value * sqrt(log_ratio_var_n0n1)

# Back-transform to get confidence intervals for the ratio
lower_ci_n0n1 <- exp(lower_log_n0n1)
upper_ci_n0n1 <- exp(upper_log_n0n1)

# Add CI to the plot
plot_data_n0n1 <- data.frame(
  time = manual_times,
  ratio = survival_ratio_n0n1,
  lower_ci = lower_ci_n0n1,
  upper_ci = upper_ci_n0n1
  
)



```

```{r}

ggplot(plot_data_n0n1, aes(x = time)) +
  # Survival ratio line
  geom_line(aes(y = ratio, color = "Survival Ratio"), size = 1.2) +
  
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  
  geom_hline(aes(yintercept = 1, linetype = "No Difference Line"), color = "red", size = 1 ) +
  
  scale_color_manual(
    name = "Line",
    values = c("Survival Ratio" = "green")
  ) +
  scale_fill_manual(
    name = "Confidence Bands",
    values = c(
      "95% Confidence Interval" = "blue"
    )
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("No Difference Line" = "dashed")
  ) +
  
  
  labs(
    title = "Survival Ratio plot for N0/N2 ",
    x = "Time (Years)",
    y = "Survival Ratio (Pathologic N0/N2)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  # Position of the legend
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )+ 
  annotate("text", x = max(plot_data_stage2_4$Time) * 0.3, y = 2, label = "Higher survival N0 ", hjust = 0) +
  annotate("text", x = max(plot_data_stage2_4$Time) * 0.3, y = 0.5, label = "Higher survival for N2", hjust = 0)  





```

#SURVIVAL DIFFERENCE

```{r}
# Compute survival difference
survival_diff_n0n1 <- surv_n0 - surv_n1

# Compute standard error for difference
diff_var_n0n1 <- var_n0 + var_n1
z_value <- qnorm(0.975)
lower_ci_n0n1_dif <- survival_diff_n0n1 - z_value * sqrt(diff_var_n0n1)
upper_ci_n0n1_dif<- survival_diff_n0n1 + z_value * sqrt(diff_var_n0n1)

# Prepare data for plotting
plot_data_diff_n0n1 <- data.frame(
  time = manual_times,
  difference = survival_diff_n0n1,
  lower_ci = lower_ci_n0n1_dif,
  upper_ci = upper_ci_n0n1_dif
)

# Generate survival difference plot
ggplot(plot_data_diff_n0n1, aes(x = time)) +
  geom_line(aes(y = difference, color = "Survival Difference"), size = 1.2) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  geom_hline(aes(yintercept = 0, linetype = "No Difference Line"), color = "red", size = 1) +
  scale_color_manual(name = "Line", values = c("Survival Difference" = "green")) +
  scale_fill_manual(name = "Confidence Bands", values = c("95% Confidence Interval" = "blue")) +
  scale_linetype_manual(name = "Reference", values = c("No Difference Line" = "dashed")) +
  labs(
    title = "Survival Difference Plot TNM N0 VS N2 ",
    x = "Time (Years)",
    y = "Survival Difference (pathologic LN N0/N2)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 18)
  )



```

############Function generating kaplan meier plot \####[TO BE CONTINUED] The code is not working

```{r}

# Function to generate survival ratio plots
generate_survival_plot <- function(data, time_col, event_col, group_col, title = "Survival Ratio Plot") {
  
  data[[group_col]] <- as.factor(data[[group_col]])

  
  surv_object <- Surv(time = data[[time_col]], event = data[[event_col]])

 
  fit <- survfit(as.formula(paste0("surv_object ~ ", group_col)), data = data)

  
  plot <- ggsurvplot(
    fit,
    data = data,
    size = 1.2,                     
    conf.int = TRUE,                 
    pval = TRUE,                     
    risk.table = TRUE,                
    ggtheme = theme_minimal(),        
    title = title,
    legend.title = group_col,
    legend.labs = levels(data[[group_col]]) 
  )
  
  return(plot)
}


filtered_data <- surv_brca_cli %>%
  filter(pathologic_stage %in% c("Stage II", "Stage III"))


generate_survival_plot(filtered_data, "time_in_years", "vital_status", "pathologic_stage", "Survival Comparison: Stage II vs Stage III")







```
