---
title: "Poster content"
author: "Victor Rwandarwacu"
date: "2025-01-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Schema

what is survival analysis with formulas? survival analysis is used when the out come is death ( other terms event history analysis,duration analysis, transition analysis, reliability analysis or failure time modelling)

time origin can be time of diagnosis or time of enrollment in the study

#### why not use statistical statistical methods

-survival data are generally not symetrically distributed (right skewed ) - censoring :when the end point has not been observed for the individual( right , left and admnistrative censoring )

assumption : actual survival time doesn' depend on any mechanism that causes censoring . indipendent or non informative censoring.

what is the difference between survival function and pdf or cdf

concepts : censoring dataand manipulation

-   **Variance Calculation**: For independent samples, the variance of the survival ratio $R(t) = \frac{S_1(t)}{S_2(t)}$ ​ is calculated using the standard errors (SEs) of the survival probabilities for each group:

    $\text{Var}(\log(R(t))) = \frac{\text{Var}(S_1(t))}S_1(t)^2} + \frac{\text{Var}(S_2(t))}{S_2(t)^2}$

    -   S1(t)S_1(t)S1​(t) and S2(t)S_2(t)S2​(t): Survival probabilities for groups 1 and 2, respectively.

    -   Var(S1(t))=(SE(S1(t)))2\text{Var}(S_1(t)) = (\text{SE}(S_1(t)))\^2Var(S1​(t))=(SE(S1​(t)))2 and Var(S2(t))=(SE(S2(t)))2\text{Var}(S_2(t)) = (\text{SE}(S_2(t)))\^2Var(S2​(t))=(SE(S2​(t)))2: Variance of survival probabilities.

-   **Confidence Interval Calculation**:

    -   Log-transform the survival ratio: log⁡(R(t))=log⁡(S1(t))−log⁡(S2(t))\log(R(t)) = \log(S_1(t)) - \log(S_2(t))log(R(t))=log(S1​(t))−log(S2​(t))

    -   Compute the 95% CI for the log-ratio: CIlog⁡=log⁡(R(t))±z⋅Var(log⁡(R(t)))\text{CI}\_{\log} = \log(R(t)) \pm z \cdot \sqrt{\text{Var}(\log(R(t)))}CIlog​=log(R(t))±z⋅Var(log(R(t)))​ where z=1.96z = 1.96z=1.96 for a 95% CI.

    -   Back-transform the CI to the original scale: CIratio=[exp⁡(CIlog⁡,lower),exp⁡(CIlog⁡,upper)]\text{CI}\_{\text{ratio}} = \left[\exp(\text{CI}_{\log, \text{lower}}), \exp(\text{CI}_{\log, \text{upper}})\right]CIratio​=[exp(CIlog,lower​),exp(CIlog,upper​)]

## Rationale for the project

There is a challenges in interpretation of survival graphs in practice

###use of survival ratio and survival difference cite papers about time to event data visualization

```{r }
# Required libraries
library(survival)
library(ggplot2)

surv_brca_cli<- read.csv("surv_brca_cli.csv")
surv_object <- Surv(time = surv_brca_cli$time_in_years, event = surv_brca_cli$vital_status)
```

## KM Plot including all covariates

```{r}
#fit the kaplan maier model for pathologic stage 
fit_path_stage <- survfit(surv_object ~ pathologic_stage , data = surv_brca_cli)

ggsurvplot(
  fit_path_stage,
  size = 1.5,                  # Line thickness
  font.x = c(16, "bold"),      # Font size for X-axis
  font.y = c(16, "bold"),      # Font size for Y-axis
  font.tickslab = c(10),       # Font size for tick labels
  legend = "right",             # Legend position
  conf.int = TRUE,                
  risk.table = TRUE,             
  pval = TRUE,                    
  xlab = "Time(years)",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),      
  legend.title = "pathological stage",     
  risk.table.height = 0.5,        
  surv.median.line = "hv" 
  
)


```

## Kaplan Meier plot for stage 2 and stage 3

```{r}
path_stage2_3_df<- surv_brca_cli|> select(time,vital_status,pathologic_stage)|> filter(pathologic_stage == "Stage II" | pathologic_stage == "Stage III")

fit23 <- survfit(Surv(time,vital_status)~pathologic_stage, data = path_stage2_3_df)
LogRank23<- survdiff(Surv(time,vital_status)~pathologic_stage, data = path_stage2_3_df)

ggsurvplot(
  fit23,
  conf.int = TRUE,                
  risk.table = TRUE,              
  xlab = "Time in days",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),
  pval = TRUE
)  
  


```

```{r}

#KM fit of the data
km_fit_path_stageII <- survfit(Surv(time= time, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_stage == "Stage II"))

km_fit_path_stageIII <- survfit(Surv(time= time, event=vital_status) ~ 1, data = subset(surv_brca_cli, pathologic_stage == "Stage III"))

####extract survival probabilities 

# Manually define time points (e.g., every 30 days)
manual_times <- seq(0, max(surv_brca_cli$time, na.rm = TRUE), by = 1)
# Extract survival probabilities for Pathologic stage 2
surv_pathII <- summary(km_fit_path_stageII, times = manual_times, extend = TRUE)$surv

# Extract survival probabilities for pathologic stage 3
surv_pathIII <- summary(km_fit_path_stageIII, times = manual_times, extend = TRUE)$surv
# Survival ratios and confidence intervals
survival_ratio <- surv_pathII / surv_pathIII

```

## Permutation Envelopes:

It Implements Monte Carlo permutation resampling to calculate confidence intervals (envelopes) for the survival ratios under the null hypothesis.

```{r}
# Permutation envelope setup (simplified for demonstration)
set.seed(1995)
n_permutations <- 1000
permuted_ratios <- matrix(NA, ncol = n_permutations, nrow = length(manual_times))

for (i in 1:n_permutations) {
  # Shuffle group labels
  surv_brca_cli$perm_pathologic_stage<- sample(surv_brca_cli$pathologic_stage)
  
  # Recompute Kaplan-Meier estimates for permuted groups
  #KM fit of the data
perm_km_fit_path_stageII <- survfit(Surv(time= time, event=vital_status) ~ 1, data = subset(surv_brca_cli, perm_pathologic_stage == "Stage II"))

perm_km_fit_path_stageIII <- survfit(Surv(time= time, event=vital_status) ~ 1, data = subset(surv_brca_cli, perm_pathologic_stage == "Stage III"))
  
 
# Extract survival probabilities for Pathologic stage 2
perm_surv_pathII <- summary( perm_km_fit_path_stageII, times = manual_times, extend = TRUE)$surv

# Extract survival probabilities for pathologic stage 3
perm_surv_pathIII <- summary(perm_km_fit_path_stageIII, times = manual_times, extend = TRUE)$surv
# Survival ratios and confidence intervals
permuted_ratios[, i] <- perm_surv_pathII / perm_surv_pathIII
  
 
}
```

```{r}


# Calculate permutation envelope
lower_bound <- apply(permuted_ratios, 1, function(x) quantile(x, 0.025, na.rm = TRUE))
upper_bound <- apply(permuted_ratios, 1, function(x) quantile(x, 0.975, na.rm = TRUE))
```

## Confidence interval estimation

```{r}

# Confidence intervals using the Delta method
# Assuming km_fit_male and km_fit_female are Kaplan-Meier survival objects

# Extract Variances  for Pathologic stage 2 and 3
var_pathII <- (summary(km_fit_path_stageII, times = manual_times, extend = TRUE)$std.err)^2
Var_pathIII <- (summary(km_fit_path_stageIII, times = manual_times, extend = TRUE)$std.err)^2



# Log-transformed survival ratio
log_ratio <- log(survival_ratio)

# Variance of the log survival ratio
log_ratio_var <- Var_pathIII / (surv_pathIII^2) + Var_pathIII / (surv_pathII^2)

# Confidence intervals for log-ratio
z_value <- qnorm(0.975)  # 1.96 for 95% confidence level
lower_log <- log_ratio - z_value * sqrt(log_ratio_var)
upper_log <- log_ratio + z_value * sqrt(log_ratio_var)

# Back-transform to get confidence intervals for the ratio
lower_ci <- exp(lower_log)
upper_ci <- exp(upper_log)

# Add CI to the plot
plot_data <- data.frame(
  time = manual_times,
  ratio = survival_ratio,
  lower_ci = lower_ci,
  upper_ci = upper_ci,
  lower_env = lower_bound,
  upper_env = upper_bound
)









```

## Plot data and ratio plot with confidence interval

```{r}
# Plot survival ratio with permutation envelope
plot_data <- data.frame(
  time = manual_times,
  ratio = survival_ratio,
   lower_ci = lower_ci,
  upper_ci = upper_ci,
  lower_env = lower_bound,
  upper_env = upper_bound
)

ggplot(plot_data, aes(x = time, y = ratio)) +
  geom_line(color = "green", size = 1.2) +

  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), fill = "blue", alpha = 0.3) +
  geom_hline(aes(yintercept = 1, linetype = "No difference line"), color = "red") +
  labs(
    title = "Survival Ratio Plot with Confidence Intervals and Permutation envelopes",
    x = "Time (Days)",
    y = "Survival Ratio (pathologic stage II / stage III)"
  ) +
  theme_minimal()+
  annotate("text", x = max(ratio_data$Time) * 0.3, y = 2, label = "Higher survival for Pathologic stage 2", hjust = 0) +
  annotate("text", x = max(ratio_data$Time) * 0.3, y = 0.5, label = "Higher survival for pathologic stage 3", hjust = 0) 





```

## Plot data and ratio plot with confidence interval and envelope

You can also embed plots, for example:

```{r , echo=FALSE}

ggplot(plot_data, aes(x = time)) +
  # Survival ratio line
  geom_line(aes(y = ratio, color = "Survival Ratio"), size = 1.2) +
  
  # Permutation envelope ribbon
  geom_ribbon(aes(ymin = lower_env, ymax = upper_env, fill = "95% Permutation Envelope"), alpha = 0.5) +
  
  # Confidence interval ribbon
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill = "95% Confidence Interval"), alpha = 0.3) +
  
  # Reference line
  geom_hline(aes(yintercept = 1, linetype = "No Difference Line"), color = "red", size = 1 ) +
  
  
  # Custom scales for the legend
  scale_color_manual(
    name = "Line",
    values = c("Survival Ratio" = "green")
  ) +
  scale_fill_manual(
    name = "Confidence Bands",
    values = c(
      "95% Permutation Envelope" = "yellow",
      "95% Confidence Interval" = "blue"
    )
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("No Difference Line" = "dashed")
  ) +
  
  # Labels and theme
  labs(
    title = "Survival Ratio Plot with Confidence Intervals and Permutation Envelopes",
    x = "Time (Days)",
    y = "Survival Ratio (Pathologic Stage II / Stage III)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",  # Position of the legend
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )+ 
  annotate("text", x = max(ratio_data$Time) * 0.3, y = 2, label = "Higher survival for Pathologic stage 2", hjust = 0) +
  annotate("text", x = max(ratio_data$Time) * 0.3, y = 0.5, label = "Higher survival for pathologic stage 3", hjust = 0) 

```
