---
title: MSc Health Data Science Project Title
author:
  - name: Victor Pacifique RWANDARWACU
  - name: Shyamsundar
    affil: 1
affiliation:
  - num: 1
    address: School of Mathematical and Statistical Sciences, University of Galway
logoleft_name: https&#58;//sport.universityofgalway.ie/assets/img/logo2.png

poster_height: "594mm"
poster_width: "841mm"
column_numbers: 4
column_margins: "0.5in"
primary_colour: "#a80050" # Secondary colour use for poster design.
secondary_colour: "#84003d" # Main colour used for poster design.
accent_colour: "#e6007e" # A third colour option.
author_textcol: "white"
title_textsize: "90pt"
author_textsize: "60pt"
body_textsize: "25pt"
output: 
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false
knit: pagedown::chrome_print
---

```{css, echo=FALSE}
div.logo_left{
  width: 20%;
}
div.poster_title{
  width: 80%;
}
.section h4 {
    break-after: column;
}
div.footnotes {
    font-size: 18pt;
}
section{
    column-count : 4 !important;
    column-gap : 0.5in !important;
}
```

<!-- Don't change anything above, except the title and author names, unless you know what you are doing. -->

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
# Load any additional libraries here
library(tidyverse)
library(survival)
library(survminer)
library(kableExtra)
```

# Background and Problem

This template has been produced using the `posterdown` package. It generates a standard conference-style poster layout.

A common problem in clinical research is messed-up data[^1]. It can lead to bias or nonsensical results. Appropriate statistical analysis can result in superb findings in such cases, with the right assumptions.

[^1]: Evans 2010, doi: 10.6030/1939-067x-3.1.1

## Objectives of Project

1.  Explain the statistical issue or research question in health data science to be investigated.
2.  Explore survival analysis using non-parametric and parametric methods.
3.  Apply methods to the lung cancer dataset and interpret results.

# Data Sources and Datasets

Our data are taken from the `survival` package in R. Specifically, we will use the `lung` dataset, which contains survival times for lung cancer patients.

#### 

# Survival Analysis and Formulas

## Relationship Between Distribution $F(t)$ and Survival Function $S(t)$:

$$
F(t)=\mathrm{P}(T<t)=\int_0^t f(u) \mathrm{d} u
$$

$$
S(t)=\mathrm{P}(T \geqslant t)=1-F(t)   
$$

## Life Table Estimate:

$$S^*(t)=\prod_{i=1}^{j-1}\left(\frac{n_i^{\prime}-d_i}{n_i^{\prime}}\right)$$

## Overall Survival Formula for Kaplan-Meier:

$$\hat{S}(t)=\prod_{j=1}^k\left(\frac{n_j-d_j}{n_j}\right)$$

####

# Early Results: Descriptive Statistics of Datasets

Summary statistics of the `lung` dataset are shown in Table \@ref(tab:lung-summary):

```{r lung-summary, out.width='80%'}
lung_summary <- lung %>%
  summarise(
    Total_Patients = n(),
    Median_Age = median(age, na.rm = TRUE),
    Male_Percentage = mean(sex == 1, na.rm = TRUE) * 100,
    Median_Survival_Time = median(time, na.rm = TRUE)
  )
knitr::kable(lung_summary, caption = 'Summary of the lung cancer dataset.', align = 'c', "html") %>%
  kable_styling(font_size = 20)
```

#### 

# Kaplan-Meier Estimate

The Kaplan-Meier survival curve is shown in Figure \@ref(fig:km-plot):

```{r km-plot, fig.align='center', fig.cap='Kaplan-Meier Survival Curve.', fig.height=5}

fit <- survfit(Surv(time, status) ~ sex, data = lung)


ggsurvplot(fit, 
           conf.int = TRUE,       
           pval = TRUE,
           risk.table = TRUE,
           xlab = "Time (days)", 
           ylab = "Survival Probability", 
           legend.title = "Sex",
           legend.labs = c("Male", "Female"),
           ggtheme = theme_minimal())
```

####

# Parametric Survival Analysis

We also fit parametric survival models (e.g., Weibull) to the lung cancer dataset. See Figure \@ref(fig:parametric-plot):

```{r parametric-plot, fig.align='center', fig.cap='Parametric Survival Analysis (Weibull).', fig.height=5}
weibull_fit <- survreg(Surv(time, status) ~ 1, data = lung, dist = "weibull")
time_seq <- seq(min(lung$time), max(lung$time), length.out = 100)
pred_survival <- predict(weibull_fit, newdata = data.frame(time = time_seq), type = "quantile", p = 0.5)
plot(time_seq, exp(-time_seq/weibull_fit$scale), type = "l", col = "blue", lwd = 2,
     xlab = "Time (days)", ylab = "Survival Probability", 
     main = "Weibull Survival Curve")
```

#### 

# Next Project Steps

We plan to conduct further analysis using:

- Additional non-parametric methods.
- Validation of parametric models.
- Incorporation of covariates (e.g., age, sex).

#### 

# GitHub

The code and datasets for this project can be viewed at our GitHub repository here: <https://github.com/>

# References

1. Evans 2010, doi: 10.6030/1939-067x-3.1.1
2. Murphy et al. 2005 doi: 10.1186/1468-6708-6-11
