---
title: "Survival Analysis of Dialysis Patients"
author: 'Shyamsundar '
date: "2025-01-25"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Schema 

### What is survival analysis?
Survival analysis is used to analyze time-to-event data, where the event could be death, relapse, or another significant milestone. Other terms include event history analysis, duration analysis, reliability analysis, or failure time modeling.

For this dataset:
- **Time origin**: Time since enrollment in the dialysis study.
- **Event**: Death (1) or censored (0).

### Why not use standard statistical methods?
1. Survival data are typically right-skewed.
2. **Censoring**: Some participants' endpoints are not observed (e.g., still alive or lost to follow-up).

**Assumption**: Censoring is non-informative, meaning it is unrelated to the likelihood of the event occurring.

### Key Concepts
- Survival function: Probability of surviving beyond a given time.
- Probability density function (PDF): Rate at which events occur.
- Censoring types: Right, left, or interval censored.

## Rationale for the Project
Visualizing survival graphs can be challenging in practice. This analysis incorporates survival ratio and survival difference plots to aid interpretation.

```{r libraries}
library(survival)
library(ggplot2)
library(dplyr)
library(survminer)

# Load the dialysis dataset
dialysis_data <- read.csv("C:/Users/shyam/OneDrive/Desktop/dialysis survival dataset.csv")

# Rename problematic column
dialysis_data <- dialysis_data %>%
  rename(time_months = `time_months`)
```

## Kaplan-Meier Survival Estimation

```{r km_fit}
# KM fit for diabetes and hypertension groups
diabetes_fit <- survfit(Surv(time = time_months, event = event) ~ disease_diabetes, data = dialysis_data)
hypert_fit <- survfit(Surv(time = time_months, event = event) ~ disease_hypert, data = dialysis_data)

# Extract time points for survival analysis
manual_times <- seq(0, max(dialysis_data$time_months, na.rm = TRUE), by = 12)

# Survival probabilities for diabetes and non-diabetes groups
diabetes_surv <- summary(diabetes_fit, times = manual_times, extend = TRUE)$surv
nondiabetes_surv <- 1 - diabetes_surv

# Survival probabilities for hypertension and non-hypertension groups
hypert_surv <- summary(hypert_fit, times = manual_times, extend = TRUE)$surv
nonhypert_surv <- 1 - hypert_surv
```

## Confidence Interval Calculation

```{r ci_calculation}
# Confidence intervals for survival probabilities
ci_diabetes <- summary(diabetes_fit, times = manual_times, extend = TRUE)$std.err
ci_hypert <- summary(hypert_fit, times = manual_times, extend = TRUE)$std.err

# Log-transformed survival ratio and its variance
log_ratio <- log(hypert_surv / diabetes_surv)
log_ratio_var <- (ci_hypert / hypert_surv)^2 + (ci_diabetes / diabetes_surv)^2

# Confidence intervals for log-ratio
z_value <- qnorm(0.975)  # 1.96 for 95% CI
lower_log <- log_ratio - z_value * sqrt(log_ratio_var)
upper_log <- log_ratio + z_value * sqrt(log_ratio_var)

# Back-transform to get confidence intervals for the ratio
lower_ci <- exp(lower_log)
upper_ci <- exp(upper_log)

# Add CI to plot data
plot_data <- data.frame(
  time = manual_times,
  survival_ratio = hypert_surv / diabetes_surv,
  survival_diff = hypert_surv - diabetes_surv,
  lower_ci = lower_ci,
  upper_ci = upper_ci
)
```

## Survival Ratio Plot

```{r survival_ratio_plot}
ggplot(plot_data, aes(x = time, y = survival_ratio)) +
  geom_line(color = "blue", size = 1.2) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), fill = "lightblue", alpha = 0.3) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(
    title = "Survival Ratio Plot (Hypertension vs Diabetes)",
    x = "Time (Months)",
    y = "Survival Ratio"
  ) +
  theme_minimal()
```

## Survival Difference Plot

```{r survival_difference_plot}
ggplot(plot_data, aes(x = time, y = survival_diff)) +
  geom_line(color = "green", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Survival Difference Plot (Hypertension vs Diabetes)",
    x = "Time (Months)",
    y = "Survival Difference"
  ) +
  theme_minimal()
```

## Risk Table

```{r risk_table}
ggsurvplot(
  list("Diabetes" = diabetes_fit, "Hypertension" = hypert_fit),
  data = dialysis_data,
  risk.table = TRUE,
  risk.table.col = "strata",
  ggtheme = theme_minimal(),
  title = "Kaplan-Meier Curves with Risk Table",
  xlab = "Time (Months)",
  ylab = "Survival Probability"
)

