---
title: "thesis"
author: "Shyamsundar Ranganath"
date: "2025-02-12"
output: html_document
---

```{r}
library(survival)
library(survminer)
library(ggplot2)
library(dplyr)
```



```{r}
dialysis_data
```

```{r}
# Plot Kaplan-Meier survival curves for each group
ggsurvplot(
  km_fit,
  data = dialysis_data,
  pval = FALSE, # Add p-value to compare groups
  conf.int = TRUE, # Add confidence intervals
  risk.table = FALSE, # Add risk table
  legend.labs = c("Non-Diabetic", "Diabetic"), # Label groups
  palette = c("blue", "red") # Assign colors to groups
)
```






```{r}
library(survival)
library(dplyr)
library(ggplot2)

# Define survival object
surv_object <- Surv(time = dialysis_data$time_years, event = dialysis_data$event)

# Fit Kaplan-Meier survival curves for the two groups
km_fit <- survfit(surv_object ~ dialysis_data$disease_diabetes, data = dialysis_data)

# Extract survival probabilities with confidence intervals
km_summary <- summary(km_fit, times = km_fit$time)

group_names <- names(km_fit$strata)
group_indices <- split(seq_along(km_summary$strata), km_summary$strata)

# Create a data frame for each group
surv_df_list <- lapply(group_names, function(group_name) {
  idx <- group_indices[[group_name]]
  data.frame(
    time = km_summary$time[idx],
    survival = km_summary$surv[idx],
    lower = km_summary$lower[idx],
    upper = km_summary$upper[idx],
    group = group_name
  )
})

# Combine the data frames into one
surv_df <- do.call(rbind, surv_df_list)

# Separate survival probabilities for the two groups
surv_0 <- surv_df %>% filter(group == group_names[1])
surv_1 <- surv_df %>% filter(group == group_names[2])

# Align survival times to compute differences
common_times <- intersect(surv_0$time, surv_1$time)
surv_0_common <- surv_0 %>% filter(time %in% common_times)
surv_1_common <- surv_1 %>% filter(time %in% common_times)

# Calculate survival difference (Group 0 - Group 1)
survival_diff <- surv_0_common$survival - surv_1_common$survival

# Compute confidence intervals for the difference
survival_diff_lower <- surv_0_common$lower - surv_1_common$upper
survival_diff_upper <- surv_0_common$upper - surv_1_common$lower

# Plot survival difference over time with confidence intervals
plot_df <- data.frame(
  time = common_times,
  survival_diff = survival_diff,
  lower = survival_diff_lower,
  upper = survival_diff_upper
)

# Plot
ggplot(plot_df, aes(x = time, y = survival_diff)) +
  geom_line(color = "blue", size = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Time (Years)",
    y = "Survival Difference (Non-Diabetic - Diabetic)",
    title = "Survival Difference Over Time with Confidence Intervals"
  ) +
  theme_minimal()

```


```{r}
set.seed(42)
n <- 100
dialysis_data <- data.frame(
  time_years = rexp(n, rate = 0.2),  # Survival times
  event = sample(c(0, 1), n, replace = TRUE, prob = c(0.3, 0.7)),  # Event occurrence
  disease_diabetes = sample(c(0, 1), n, replace = TRUE, prob = c(0.6, 0.4))  # 0 = Non-Diabetic, 1 = Diabetic
)

# Create the survival object
surv_obj <- Surv(time = dialysis_data$time_years, event = dialysis_data$event)

# Fit Kaplan-Meier survival model by diabetes status
km_fit <- survfit(surv_obj ~ disease_diabetes, data = dialysis_data)

# Plot Kaplan-Meier survival curves
ggsurvplot(
  km_fit,
  data = dialysis_data,
  conf.int = TRUE, 
  risk.table = FALSE, 
  pval = FALSE, 
  legend.labs = c("Non-Diabetic", "Diabetic"), 
  palette = c("blue", "red"), 
  xlab = "Time (Years)",
  ylab = "Survival Probability",
  ggtheme = theme_minimal()
)
```

The red curve which represents the diabetic patients tends to decline faster than the blue curve which is represented by the non diabetic patients. This suggests that diabetic patients have a lower survival probability over time compared to the non diabetic patients.

The cross marks(+) on the lined indicate censored observations meaning that the follow up ended before an event occured for those induviduals.

The blue region is narrower whiich suggest more stable estimate compared to diabetics.
```{r}
library(viridis)
```

```{r}
set.seed(42)
n <- 100
dialysis_data <- data.frame(
  time_years = rexp(n, rate = 0.2),  
  event = sample(c(0, 1), n, replace = TRUE, prob = c(0.3, 0.7)),  
  disease_diabetes = sample(c(0, 1), n, replace = TRUE, prob = c(0.6, 0.4))  # 0 = Non-Diabetic, 1 = Diabetic
)

# Kaplan-Meier model
surv_obj <- Surv(time = dialysis_data$time_years, event = dialysis_data$event)
km_fit <- survfit(surv_obj ~ disease_diabetes, data = dialysis_data)

# Extract survival probabilities
surv_data <- data.frame(
  time = km_fit$time,
  survival = km_fit$surv,
  group = rep(c("Non-Diabetic", "Diabetic"), each = length(km_fit$time) / 2)
)

# Create a contour plot using ggplot2
ggplot(surv_data, aes(x = time, y = survival, z = survival, fill = ..level..)) +
  geom_density_2d_filled() +  
  scale_fill_viridis_d() +  
  labs(title = "Survival Probability Contour Plot",
       x = "Time (Years)",
       y = "Survival Probability") +
  theme_minimal()
```

Yellow and green areas indicate higher concentrations of survival probabilities while Dark Blue and pruple areas indicate lower concentrations of survival probabilities.

From the above plot most survival probability estimates cluster within the first 5 years. After 5-10 years the probabilities decrease. After 10 years there are few survival data points.

Most patients experience a decline in survival probability within the first few years.Survival beyond 10 years is rare.


