---
title: "Survey_analysis"
author: "Victor Rwandarwacu"
date: "2025-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## survey Analysis 

```{r}
#library 
library(readxl)
library(janitor)
library (tidyverse)
library(table1)
library(rpart)
library(rpart.plot)




```


```{r}
#import data 
survey_data <- read_xlsx("Survey_coded_export20250521.xlsx")


```
#Tidy data & rename columns 

```{r}

survey_data <- survey_data %>% clean_names()
survey_data <- survey_data %>% 
  select(-pre_km 
,-pre_survdiff,-pre_mrl,-pre_mr_ldiff,-post_km,-post_survdiff,-post_mrl,-post_mr_ldiff)



sapply(survey_data,class)
#colSums(is.na(survey_data))

colnames(survey_data)

categorical_vars <- c("qualification","country","pre_km_code","pre_survdiff_code","pre_mrl_code",       "pre_mr_ldiff_code","post_km_code","post_survdiff_code","post_mrl_code","post_mr_ldiff_code","ranking1_km","ranking1_survdiff","ranking1_mrl","ranking1_mr_ldiff","ranking2_km","ranking2_survdiff","ranking2_mrl","ranking2_mr_ldiff")

survey_data[categorical_vars] <- lapply(survey_data[categorical_vars],factor)
summary(survey_data)
```



## table 1_ demographic summary 

```{r}
table1(~ country+qualification+knowledge_rate , data = survey_data, caption = "<b>Descriptive summary" )
```

## Exploratory visualisation 

## self rated knowledge score 
```{r}
score_rate_distribution <- survey_data %>%  group_by(knowledge_rate) %>% summarise( count=n())
score_rate_distribution

distr_knowledge_score<- ggplot(data = score_rate_distribution, aes(x = as.factor(knowledge_rate), y = count)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  geom_text(aes(label = count), vjust = -0.3, size = 4.5) +
  labs(title = "Distribution of Self-rated Knowledge Scores",
       x = "Self-rated Knowledge Score",
       y = NULL) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.title = element_text(face = "bold", size = 16),         
    axis.title.x = element_text(size = 14, face = "bold"),       
    axis.text.x = element_text(size = 12)
  )

ggsave("Distribution of Self-rated Knowledge Scores.png", plot = distr_knowledge_score, width = 10, height = 6, dpi = 300,bg="white")

distr_knowledge_score
```

## Do those who said they know actually know?
```{r}

# Create a proportion of correct answers on pre-test
survey_data <- survey_data %>%
  mutate(
    pre_correct = rowMeans(across(ends_with("pre_km_code") | 
                                  ends_with("pre_survdiff_code") | 
                                  ends_with("pre_mrl_code") | 
                                  ends_with("pre_mr_ldiff_code"), 
                                  ~ . == "CORRECT", 
                                  .names = "correct_{col}"), 
                           na.rm = TRUE)
  )

# Average correctness by knowledge level
knowledge_plot_data <- survey_data %>%
  group_by(knowledge_rate) %>%
  summarise(avg_pre_correct = mean(pre_correct, na.rm = TRUE))

# Plot
avg_prop_pretest <- ggplot(knowledge_plot_data, aes(x = factor(knowledge_rate), y = avg_pre_correct)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Proportion of Correct Interpretation per Self-rated Knowledge",
    x = "Self-rated Knowledge (1 = Low, 10 = High)",
    y = "Average Proportion Correct (Pre-Test)"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(face = "bold", size = 16),         
    axis.title.x = element_text(size = 14, face = "bold"),       
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 14, face = "bold"),       
    axis.text.y = element_text(size = 12)
  )

ggsave("Average Proportion Correct (Pre-Test).png", plot = avg_prop_pretest, width = 10, height = 6, dpi = 300,bg="white")
avg_prop_pretest
```


# Question 2: Did those who don’t know learn (pre vs post)?
```{r}


# Create total pre/post correctness
survey_data <- survey_data %>%
  mutate(
    pre_score = rowSums(across(ends_with("pre_km_code") | 
                               ends_with("pre_survdiff_code") | 
                               ends_with("pre_mrl_code") | 
                               ends_with("pre_mr_ldiff_code"), ~ . == "CORRECT")),
    post_score = rowSums(across(ends_with("post_km_code") | 
                                ends_with("post_survdiff_code") | 
                                ends_with("post_mrl_code") | 
                                ends_with("post_mr_ldiff_code"), ~ . == "CORRECT")),
    learning_gain = post_score - pre_score
  )

# Filter for those with low knowledge rate (1 or 5)
low_knowledge_learners <- survey_data %>%
  filter(knowledge_rate <= 5)

# Plot learning gain
knowledge_gain <- ggplot(low_knowledge_learners, aes(x = factor(knowledge_rate), y = learning_gain)) +
  geom_boxplot(fill = "lightgreen") +
  labs(
    title = "Learning Gain for Low-Knowledge Respondents",
    x = "Self-rated Knowledge (Low)",
    y = "Change in Correct Answers (Post - Pre)"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(face = "bold", size = 16),         
    axis.title.x = element_text(size = 14, face = "bold"),       
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 14, face = "bold"),       
    axis.text.y = element_text(size = 12)
  )

ggsave("Learning Gain for Low-Knowledge Respondents.png", plot = knowledge_gain, width = 10, height = 6, dpi = 300,bg="white")

knowledge_gain
```

## knowledge gain faceted by types of plots ( pre versus post )
```{r}

# Prepare long-format data for plotting
correctness_long <- survey_data %>%
  select(ends_with("code")) %>%
  mutate(id = row_number()) %>%
  pivot_longer(
    cols = -id,
    names_to = "question_full",
    values_to = "correctness"
  ) %>%
  mutate(
    phase = factor(
      case_when(
      str_detect(question_full, "^pre_") ~ "Pre_Learning",
      str_detect(question_full, "^post_") ~ "Post_Learning"
    ),levels = c("Pre_Learning", "Post_Learning")  # Specify order explicitly
),
    question = case_when(
      str_detect(question_full, "km") ~ "Kaplan Meier",
      str_detect(question_full, "survdiff") ~ "Survival difference ",
      str_detect(question_full, "mrl") ~ "MRL",
      str_detect(question_full, "mr_ldiff") ~ "MRL Difference"
    ),
    correctness = factor(correctness, levels = c("INCORRECT", "CORRECT")) # Ensure CORRECT is on the bottom
  )

# stacked proportional bars, faceted by question and phase
prop_correct_pre_vs_post <- ggplot(correctness_long, aes(x = 1, fill = correctness)) +
  geom_bar(position = "fill", width = 0.5) +
  facet_grid(rows = vars(question), cols = vars(phase), switch = "y") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("CORRECT" = "#0072B2", "INCORRECT" = "#E69F00")) +
  labs(
    title = "Proportion of Correct vs Incorrect Interpretations(Pre vs Post)",
    x = NULL,
    y = "Proportion",
    fill = "Answer"
  ) +
  theme_minimal() +
  theme(
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank(),
  strip.text.y.left = element_text(angle = 0, size = 12,face = "bold"),
  strip.placement = "outside",
  strip.text.x = element_text(size = 10,face = "bold"),
  plot.title = element_text(face = "bold", size = 14),
  axis.title.x = element_text(size = 14, face = "bold"),       
  axis.title.y = element_text(size = 14, face = "bold"),       
  axis.text.y = element_text(size = 8,face = "bold"),
  legend.position = "top"
  
)

ggsave("Proportion of Correct vs Incorrect (Pre vs Post).png", plot = prop_correct_pre_vs_post, width = 10, height = 6, dpi = 300,bg="white")
prop_correct_pre_vs_post
```

## McNemar test for each question 
```{r}
library(dplyr)
library(tidyr)
library(broom)

# Convert to wide format
correctness_wide <- correctness_long %>%
  select(id, question, phase, correctness) %>%
  pivot_wider(
    names_from = phase,
    values_from = correctness
  )

# Run McNemar test for each question
mcnemar_results <- correctness_wide %>%
  group_by(question) %>%
  summarise(
    p_value = broom::tidy(
      stats::mcnemar.test(
        table(Pre_Learning, Post_Learning)
      )
    )$p.value
  )

print(mcnemar_results)

```


## side by side bar chart 

```{r}


# Reshape to long format with proper labels for Pre and Post
ranking_bar <- survey_data %>%
  pivot_longer(
    cols = starts_with("ranking"),
    names_to = "ranking_type",
    values_to = "rank"
  ) %>%
  mutate(
    question = case_when(
      str_detect(ranking_type, "km") ~ "Kaplan Meier(KM)",
      str_detect(ranking_type, "survdiff") ~ "Survival difference",
      str_detect(ranking_type, "mrl$") ~ "Mean Residual Life(MRL)",
      str_detect(ranking_type, "mr_ldiff") ~ "MRL difference"
    ),
    preference = case_when(
      str_detect(ranking_type, "ranking1") ~ "Easy to understand",
      str_detect(ranking_type, "ranking2") ~ "use with patient"
    ),
    rank = factor(rank, levels = c("1", "2", "3", "4"))  # Ensure correct rank order
  ) %>%
  drop_na(rank)

#  Side-by-side bars for Pre vs Post,
ranking_bar_plot <- ggplot(ranking_bar, aes(x = rank, fill = preference)) +
  geom_bar(position = "dodge") +
  facet_wrap(~question) +
  scale_fill_manual(values = c("Easy to understand" = "#0072B2", "use with patient" = "#E69F00")) +
  labs(
    title = "Distribution of Rankings ",
    x = "Rank",
    y = "Number of Participants",
    fill = "preference"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16),         
    axis.title.x = element_text(size = 14, face = "bold"),       
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 14, face = "bold"),       
    axis.text.y = element_text(size = 12),
    strip.text = element_text(size = 12, face = "bold")
   
)

ggsave("Distribution of Rankings.png", plot = ranking_bar_plot, width = 10, height = 6, dpi = 300,bg="white")
ranking_bar_plot
```



#scatter plot with jitter 


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Prepare the data with factors
ranking_scatter <- survey_data %>%
  select(id,
         ranking1_km, ranking2_km,
         ranking1_survdiff, ranking2_survdiff,
         ranking1_mrl, ranking2_mrl,
         ranking1_mr_ldiff, ranking2_mr_ldiff) %>%
  pivot_longer(
    cols = -id,
    names_to = c("phase", "question"),
    names_pattern = "ranking(\\d)_(.*)",
    values_to = "rank"
  ) %>%
  mutate(
    phase = recode(phase, `1` = "Pre", `2` = "Post"),
    question = case_when(
      question == "km" ~ "Kaplan Meier",
      question == "survdiff" ~ "Survival difference",
      question == "mrl" ~ "MRL",
      question == "mr_ldiff" ~ "MRL difference",
      TRUE ~ question
    )
  ) %>%
  pivot_wider(names_from = phase, values_from = rank) %>%
  drop_na() %>%
  mutate(
    Pre = factor(Pre, levels = c("1", "2", "3", "4")),
    Post = factor(Post, levels = c("1", "2", "3", "4"))
  )

```


```{r}


ranking_scatter_plot <-ggplot(ranking_scatter, aes(x = Pre, y = Post)) +
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.4, color = "#0072B2", size = 2.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~question) +
 scale_x_discrete(name = "Easy to understand by clinician") +
  scale_y_discrete(name = "Easy to use in consultation") +
  labs(
    title = "Agreement Between Rankings"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 12),
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold"),       
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 14, face = "bold"),       
    axis.text.y = element_text(size = 12),
  )
ggsave(" Agreement of Rankings.png", plot = ranking_scatter_plot, width = 10, height = 6, dpi = 300,bg="white")
ranking_scatter_plot

```




```{r}

# Make sure Pre and Post are factors with defined levels
ranking_scatter <- ranking_scatter %>%
  mutate(
    Pre = factor(Pre, levels = 1:4),
    Post = factor(Post, levels = 1:4)
  )

# Count occurrences of each Pre-Post combination
ranking_counts <- ranking_scatter %>%
  count(question, Pre, Post)

# Plot
g<- ggplot(ranking_counts, aes(x = Pre, y = Post)) +
  geom_jitter(data = ranking_scatter, aes(x = Pre, y = Post), 
              width = 0.2, height = 0.2, alpha = 0.3, color = "#0072B2", size = 2.5) +
  geom_text(aes(label = n), color = "black", fontface = "bold", size = 4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~question) +
  scale_x_discrete(name = "Easy to understand by clinician") +
  scale_y_discrete(name = "Easy to use in consultation") +
  labs(
    title = "Agreement Between Rankings (with Counts)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 12)
  )


ggplotly(g)

```



## Ranking chnages (allevial plot)
```{r}

library(ggalluvial)


# Create long format for rankings
ranking_long <- survey_data %>%
  mutate(id = row_number()) %>%
  pivot_longer(
    cols = starts_with("ranking"),
    names_to = "ranking_type",
    values_to = "rank"
  ) %>%
  mutate(
    plot_type = case_when(
      str_detect(ranking_type, "km") ~ "Kaplan Meier(KM)",
      str_detect(ranking_type, "survdiff") ~ "Survival difference",
      str_detect(ranking_type, "mrl$") ~ "Mean Residual Life(MRL)",
      str_detect(ranking_type, "mr_ldiff") ~ "MRL difference"
    ),
    preference = case_when(
      str_detect(ranking_type, "ranking1") ~ "Pre",
      str_detect(ranking_type, "ranking2") ~ "Post"
    )
  ) %>%
  drop_na(rank)

# Reshape into wide format to define flows: one row per person-question
ranking_wide <- ranking_long %>%
  pivot_wider(id_cols = c(id, plot_type),
              names_from = preference,
              values_from = rank)

# Plot: Alluvial diagram
ranking_change_plot <- ggplot(ranking_wide,
       aes(axis1 = as.factor(Pre),
           axis2 = as.factor(Post))) +
  geom_alluvium(aes(fill = plot_type), width = 0.25, alpha = 0.8) +
  geom_stratum(width = 0.25, color = "black", fill = "gray90") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  facet_wrap(~plot_type) +
  scale_x_discrete(limits = c("Easy to understand", "use with patient")) +
  labs(
    title = "Ranking Changes: Easy to understand by clinician versus Use with patient in consultation",
    x = "preference", y = "Number of Participants"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(face = "bold", size = 14))

ggsave("Ranking change.png", plot = ranking_change_plot, width = 10, height = 6, dpi = 300,bg="white")
```



## Fitting a tree to the data set #####################################################


```{r}

library(rpart.plot)
# Create useful predictors
tree_data <- survey_data %>%
  mutate(
    pre_total_correct = rowSums(across(c(pre_km_code, pre_survdiff_code, pre_mrl_code, pre_mr_ldiff_code), ~ . == "CORRECT")),
    post_mrl_correct = factor(post_mrl_code, levels = c("INCORRECT", "CORRECT")),
    qualification = factor(qualification),
    Self_rated_knowledge = as.numeric(knowledge_rate)
  ) %>%
  select(post_mrl_correct, Self_rated_knowledge, pre_total_correct, qualification) %>%
  drop_na()

tree_data <- tree_data %>%
  mutate(
    qualification = case_when(
      qualification == "Medical Doctor (Specialist or postgraduate)" ~ "MD_Specialist",
      qualification == "Medical Doctor (General Practitioner)" ~ "MD_GP",
      qualification == "Medical student" ~ "Med_student"
    )
  )
head(tree_data)
# Fit decision tree model
tree_model <- rpart(post_mrl_correct ~ ., data = tree_data, method = "class", control = rpart.control(cp = 0, minsplit = 10,minbucket =5 ,maxdepth = 10))

# Plot the tree
rpart.plot(tree_model, type = 4, extra = 104,
           main = "Prediction of correct MRL plot interpretation post learning section")

```
## predict learning gain 

```{r}


# data process
tree_data_gain <- survey_data %>%
  mutate(
    # Convert correctness codes to numeric (1 = correct, 0 = incorrect)
    pre_total_correct = rowSums(across(c(pre_km_code, pre_survdiff_code, pre_mrl_code, pre_mr_ldiff_code), ~ . == "CORRECT")),
    post_total_correct = rowSums(across(c(post_km_code, post_survdiff_code, post_mrl_code, post_mr_ldiff_code), ~ . == "CORRECT")),
    learning_gain = post_total_correct - pre_total_correct,
    qualification = case_when(
      qualification == "Medical Doctor (Specialist or postgraduate)" ~ "MD_Specialist",
      qualification == "Medical Doctor (General Practitioner)" ~ "MD_GP",
      qualification == "Medical student" ~ "Med_student"
    ),
    knowledge_rate = as.numeric(knowledge_rate)
  ) %>%
  select(learning_gain, knowledge_rate, pre_total_correct, qualification) %>%
  drop_na()

# Fit regression tree model
reg_tree <- rpart(learning_gain ~ ., data = tree_data_gain, method = "anova", control = rpart.control(cp = 0, minsplit = 10,minbucket =5 ,maxdepth = 10))

# Plot the regression tree
regression_tree_gain <- rpart.plot(reg_tree, type = 4, extra = 101,
           main = "Regression Tree Predicting Learning Gain")

ggsave("Regression Tree Predicting Learning Gain.png", plot = regression_tree_gain, width = 10, height = 6, dpi = 300,bg="white")

```

