---
title: "Survey_analysis"
author: "Victor Rwandarwacu"
date: "2025-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## survey Analysis 
dggfhfdhd

```{r}
#library 
library(readxl)
library(janitor)
library (tidyverse)
library(table1)
library(rpart)
library(rpart.plot)




```


```{r}
#import data 
survey_data <- read_xlsx("Survey_coded_export20250521.xlsx")


```
#Tidy data & rename columns 

```{r}

survey_data <- survey_data %>% clean_names()
survey_data <- survey_data %>% 
  select(-pre_km 
,-pre_survdiff,-pre_mrl,-pre_mr_ldiff,-post_km,-post_survdiff,-post_mrl,-post_mr_ldiff)



sapply(survey_data,class)
#colSums(is.na(survey_data))

colnames(survey_data)

categorical_vars <- c("qualification","country","pre_km_code","pre_survdiff_code","pre_mrl_code",       "pre_mr_ldiff_code","post_km_code","post_survdiff_code","post_mrl_code","post_mr_ldiff_code","ranking1_km","ranking1_survdiff","ranking1_mrl","ranking1_mr_ldiff","ranking2_km","ranking2_survdiff","ranking2_mrl","ranking2_mr_ldiff")

survey_data[categorical_vars] <- lapply(survey_data[categorical_vars],factor)
summary(survey_data)
```



## table 1_ demographic summary 

```{r}
table1(~ country+qualification+knowledge_rate , data = survey_data, caption = "<b>Descriptive summary" )
```

## Exploratory visualisation 

## self rated knowledge score 
```{r}
score_rate_distribution <- survey_data %>%  group_by(knowledge_rate) %>% summarise( count=n())
score_rate_distribution

distr_knowledge_score<- ggplot(data = score_rate_distribution, aes(x = as.factor(knowledge_rate), y = count)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  geom_text(aes(label = count), vjust = -0.3, size = 4.5) +
  labs(title = "Distribution of Self-rated Knowledge Scores",
       x = "Self-rated Knowledge Score",
       y = NULL) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.title = element_text(face = "bold", size = 16),         
    axis.title.x = element_text(size = 14, face = "bold"),       
    axis.text.x = element_text(size = 12)
  )

ggsave("Distribution of Self-rated Knowledge Scores.png", plot = distr_knowledge_score, width = 10, height = 6, dpi = 300,bg="white")

distr_knowledge_score
```

## Do those who said they know actually know?
```{r}

# Create a proportion of correct answers on pre-test
survey_data <- survey_data %>%
  mutate(
    pre_correct = rowMeans(across(ends_with("pre_km_code") | 
                                  ends_with("pre_survdiff_code") | 
                                  ends_with("pre_mrl_code") | 
                                  ends_with("pre_mr_ldiff_code"), 
                                  ~ . == "CORRECT", 
                                  .names = "correct_{col}"), 
                           na.rm = TRUE),
     post_correct = rowMeans(across(ends_with("pre_km_code") | 
                                  ends_with("pre_survdiff_code") | 
                                  ends_with("pre_mrl_code") | 
                                  ends_with("pre_mr_ldiff_code"), 
                                  ~ . == "CORRECT", 
                                  .names = "correct_{col}"), 
                           na.rm = TRUE)
  )

# Average correctness by knowledge level
knowledge_plot_data <- survey_data %>%
  group_by(knowledge_rate) %>%
  summarise(avg_pre_correct = mean(pre_correct, na.rm = TRUE))

# Plot
avg_prop_pretest <- ggplot(knowledge_plot_data, aes(x = factor(knowledge_rate), y = avg_pre_correct)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Proportion of Correct Interpretation per Self-rated Knowledge",
    x = "Self-rated Knowledge (1 = Low, 10 = High)",
    y = "Average Proportion Correct (Pre-Test)"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(face = "bold", size = 16),         
    axis.title.x = element_text(size = 14, face = "bold"),       
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 14, face = "bold"),       
    axis.text.y = element_text(size = 12)
  )

ggsave("Average Proportion Correct (Pre-Test).png", plot = avg_prop_pretest, width = 10, height = 6, dpi = 300,bg="white")
avg_prop_pretest
```


# plot 2: 
```{r}


# Create total pre/post correctness
survey_data <- survey_data %>%
  mutate(
    pre_score = rowSums(across(ends_with("pre_km_code") | 
                               ends_with("pre_survdiff_code") | 
                               ends_with("pre_mrl_code") | 
                               ends_with("pre_mr_ldiff_code"), ~ . == "CORRECT")),
    post_score = rowSums(across(ends_with("post_km_code") | 
                                ends_with("post_survdiff_code") | 
                                ends_with("post_mrl_code") | 
                                ends_with("post_mr_ldiff_code"), ~ . == "CORRECT")),
    learning_gain = post_score - pre_score
  )

# Filter for those with low knowledge rate (1 or 5)
low_knowledge_learners <- survey_data %>%
  filter(knowledge_rate <= 5)

# Plot learning gain
knowledge_gain <- ggplot(low_knowledge_learners, aes(x = factor(knowledge_rate), y = learning_gain)) +
  geom_boxplot(fill = "lightgreen") +
  labs(
    title = "Learning Gain for Low-Knowledge Respondents",
    x = "Self-rated Knowledge (Low)",
    y = "Change in Correct Answers (Post - Pre)"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(face = "bold", size = 16),         
    axis.title.x = element_text(size = 14, face = "bold"),       
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 14, face = "bold"),       
    axis.text.y = element_text(size = 12)
  )

ggsave("Learning Gain for Low-Knowledge Respondents.png", plot = knowledge_gain, width = 10, height = 6, dpi = 300,bg="white")

knowledge_gain
```

## knowledge gain faceted by types of plots ( pre versus post )
```{r}

# Prepare long-format data for plotting
correctness_long <- survey_data %>%
  select(ends_with("code"),knowledge_rate) %>%
  mutate(id = row_number()) %>%
  pivot_longer(
    cols = -c(id,knowledge_rate),
    names_to = "plot_full",
    values_to = "correctness"
  ) %>%
  mutate(
    phase = factor(
      case_when(
      str_detect(plot_full, "^pre_") ~ "Pre_Learning",
      str_detect(plot_full, "^post_") ~ "Post_Learning"
    ),levels = c("Pre_Learning", "Post_Learning")  
),
    plot = case_when(
      str_detect(plot_full, "km") ~ "Kaplan Meier",
      str_detect(plot_full, "survdiff") ~ "Difference in Survival",
      str_detect(plot_full, "mrl") ~ "MRL",
      str_detect(plot_full, "mr_ldiff") ~ "Difference in MRL"
    ),
    correctness = factor(correctness, levels = c("INCORRECT", "CORRECT")) # Ensure CORRECT is on the bottom
  )

# stacked proportional bars, faceted by plot and phase
if (FALSE) {prop_correct_pre_vs_post <- ggplot(correctness_long, aes(x = 1, fill = correctness)) +
  geom_bar(position = "fill", width = 0.5) +
  facet_grid(rows = vars(plot), cols = vars(phase), switch = "y") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("CORRECT" = "#0072B2", "INCORRECT" = "#E69F00")) +
  labs(
    title = "Proportion of Correct vs Incorrect Interpretations(Pre vs Post)",
    x = NULL,
    y = "Proportion",
    fill = "Answer"
  ) +
  theme_minimal() +
  theme(
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank(),
  strip.text.y.left = element_text(angle = 0, size = 12,face = "bold"),
  strip.placement = "outside",
  strip.text.x = element_text(size = 10,face = "bold"),
  plot.title = element_text(face = "bold", size = 14),
  axis.title.x = element_text(size = 14, face = "bold"),       
  axis.title.y = element_text(size = 14, face = "bold"),       
  axis.text.y = element_text(size = 8,face = "bold"),
  legend.position = "top"
  
)

ggsave("Proportion of Correct vs Incorrect (Pre vs Post).png", plot = prop_correct_pre_vs_post, width = 10, height = 6, dpi = 300,bg="white")
prop_correct_pre_vs_post}
```
## overall accuracy (pre versus post )

```{r}
# Calculate overall proportion of CORRECT per phase (ignore plot type)
overall_correct_prop <- correctness_long %>%
  group_by(phase) %>%
  summarise(prop_correct = mean(correctness == "CORRECT")) %>%
  ungroup()

# Overall Pre vs Post plot
overall_bar_correct <- ggplot(overall_correct_prop, aes(x = phase, y = prop_correct, fill = phase)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_fill_manual(values = c("Pre_Learning" = "#E69F00", "Post_Learning" = "#0072B2")) +
  labs(
    title = "Overall Proportion of Correct Interpretations (Pre vs Post)",
    x = "Phase",
    y = "Proportion Correct",
    fill = "Phase"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "none"  
  )

# Print plot
overall_bar_correct
```




## grouped bar chart of the plot 
```{r}


# Calculation of  proportion of CORRECT per plot  and phase
correct_prop <- correctness_long %>%
  group_by( plot, phase) %>%
  summarise(prop_correct = mean(correctness == "CORRECT")) %>%
  ungroup()

# Grouped bar chart
grouped_bar_correct <- ggplot(correct_prop, aes(x = plot, y = prop_correct, fill = phase)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_fill_manual(values = c("Pre_Learning" = "#E69F00", "Post_Learning" = "#0072B2")) +
  labs(
    title = "Proportion of Correct Interpretations (Pre vs Post Learning) by plot",
    x = "Plot",
    y = "Proportion Correct",
    fill = "Phase"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, face = "bold", angle = 20, hjust = 1),
    axis.text.y = element_text(size = 10, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "top"
  )

# Print plot
grouped_bar_correct
```
## sigificance test for the overall learning effect 

```{r}
# Prepare per-respondent % correct pre and post
respondent_correct_prop <- correctness_long %>%
  group_by(id, phase) %>%
  summarise(prop_correct = mean(correctness == "CORRECT", .groups = "drop")) %>%
  pivot_wider(names_from = phase, values_from = prop_correct)

# Paired t-test
t_test_result <- t.test(respondent_correct_prop$Pre_Learning, respondent_correct_prop$Post_Learning, paired = TRUE)

# Wilcoxon signed-rank test (non-parametric)
wilcox_result <- wilcox.test(respondent_correct_prop$Pre_Learning, respondent_correct_prop$Post_Learning, paired = TRUE)

# Output results
t_test_result
wilcox_result
```
Wilcoxon is the safest and robust test because it does not assume normal distribution.


## McNemar test for each plot 
```{r}
library(dplyr)
library(tidyr)
library(broom)

# Convert to wide format
correctness_wide <- correctness_long %>%
  select(id, plot, phase, correctness) %>%
  pivot_wider(
    names_from = phase,
    values_from = correctness
  )

# Run McNemar test for each plot
mcnemar_results <- correctness_wide %>%
  group_by(plot) %>%
  summarise(
    p_value = broom::tidy(
      stats::mcnemar.test(
        table(Pre_Learning, Post_Learning)
      )
    )$p.value
  )

print(mcnemar_results)

```


```{r}
# Reshape to long format
respondent_correct_long <- respondent_correct_prop %>%
  pivot_longer(cols = c(Pre_Learning, Post_Learning),
               names_to = "phase",
               values_to = "prop_correct")%>%
  mutate(phase = factor(phase, levels = c("Pre_Learning", "Post_Learning")))

if (FALSE) {# Line plot
paired_line_plot <- ggplot(respondent_correct_long, aes(x = phase, y = prop_correct, group = id)) +
  geom_line(color = "gray60", size = 0.8) +  # lines for each respondent
  geom_point(size = 3, aes(color = phase)) +  # points Pre and Post
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_color_manual(values = c("Pre_Learning" = "#E69F00", "Post_Learning" = "#0072B2")) +
  labs(
    title = "Individual Changes in Correct Interpretations (Pre vs Post)",
    x = "Phase",
    y = "Proportion Correct"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "none"
  )

# Print plot
paired_line_plot}

```


```{r}

# Reuse respondent_correct_long table

# Bar plot with CI
overall_bar_CI <- ggplot(respondent_correct_long, aes(x = phase, y = prop_correct, fill = phase)) +
  stat_summary(fun = mean, geom = "bar", position = position_dodge(width = 0.6), width = 0.6) +
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.2, position = position_dodge(width = 0.6)) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_fill_manual(values = c("Pre_Learning" = "#E69F00", "Post_Learning" = "#0072B2")) +
  labs(
    title = "Overall Proportion of Correct Interpretations (with 95% CI)",
    x = "Phase",
    y = "Proportion Correct",
    fill = "Phase"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "none"
  )

# Print plot
overall_bar_CI



```


## paired line faceted by plot 

```{r}

# % correct per respondent per phase per plot
respondent_plot_correct <- correctness_long %>%
  group_by(id, plot, phase) %>%
  summarise(prop_correct = mean(correctness == "CORRECT"), .groups = "drop")

# Explicit phase ordering
respondent_plot_correct <- respondent_plot_correct %>%
  mutate(phase = factor(phase, levels = c("Pre_Learning", "Post_Learning")))


```
















## side by side bar chart Ranking plots 

```{r}


# Reshape to long format with proper labels for Pre and Post
ranking_bar <- survey_data %>%
  pivot_longer(
    cols = starts_with("ranking"),
    names_to = "ranking_type",
    values_to = "rank"
  ) %>%
  mutate(
    plot = case_when(
      str_detect(ranking_type, "km") ~ "Kaplan Meier(KM)",
      str_detect(ranking_type, "survdiff") ~ "Survival difference",
      str_detect(ranking_type, "mrl$") ~ "Mean Residual Life(MRL)",
      str_detect(ranking_type, "mr_ldiff") ~ "MRL difference"
    ),
    preference = case_when(
      str_detect(ranking_type, "ranking1") ~ "Easy to understand by clinician",
      str_detect(ranking_type, "ranking2") ~ "Easy to use in consultation"
    ),
    rank = factor(rank, levels = c("1", "2", "3", "4"))  # Ensure correct rank order
  ) %>%
  drop_na(rank)

#  Side-by-side bars for Pre vs Post,
ranking_bar_plot <- ggplot(ranking_bar, aes(x = rank, fill = preference)) +
  geom_bar(position = "dodge") +
  facet_wrap(~plot) +
  scale_fill_manual(values = c("Easy to understand by clinician" = "#0072B2", "Easy to use in consultation" = "#E69F00")) +
  labs(
    title = "Distribution of Rankings ",
    x = "Rank",
    y = "Number of Participants",
    fill = "preference"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16),         
    axis.title.x = element_text(size = 14, face = "bold"),       
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 14, face = "bold"),       
    axis.text.y = element_text(size = 12),
    strip.text = element_text(size = 12, face = "bold")
   
)

ggsave("Distribution of Rankings.png", plot = ranking_bar_plot, width = 10, height = 6, dpi = 300,bg="white")
ranking_bar_plot
```



#scatter plot with jitter 


```{r}


# Prepare the data with factors
ranking_agreement <- survey_data %>%
  select(id,
         ranking1_km, ranking2_km,
         ranking1_survdiff, ranking2_survdiff,
         ranking1_mrl, ranking2_mrl,
         ranking1_mr_ldiff, ranking2_mr_ldiff) %>%
  pivot_longer(
    cols = -id,
    names_to = c("Preference", "plot"),
    names_pattern = "ranking(\\d)_(.*)",
    values_to = "rank"
  ) %>%
  mutate(
    Preference = recode(Preference, `1` = "Clinicians_Understanding", `2` = "Use_with_patient"),
    plot = case_when(
      plot == "km" ~ "Kaplan Meier(KM)",
      plot == "survdiff" ~ "Difference in Survival",
      plot == "mrl" ~ "Mean Residual Life(MRL)",
      plot == "mr_ldiff" ~ "Difference in MRL",
      TRUE ~ plot
    )
  ) %>%
  pivot_wider(names_from = Preference, values_from = rank) %>%
  drop_na() %>%
  mutate(
  Clinicians_Understanding = factor(Clinicians_Understanding, levels = c("1", "2", "3", "4")),
  Use_with_patient = factor(Use_with_patient, levels = c("1", "2", "3", "4"))
  )

#write.csv(ranking_agreement, "ranking_agreement.csv",row.names = FALSE)
```


```{r}


ranking_agreement_plot <-ggplot(ranking_agreement, aes(x = Clinicians_Understanding, y = Use_with_patient)) +
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.4, color = "#0072B2", size = 2.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~plot) +
 scale_x_discrete(name = "Easy to understand by clinician") +
  scale_y_discrete(name = "Easy to use in consultation") +
  labs(
    title = "Agreement Between Rankings"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 12),
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold"),       
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 14, face = "bold"),       
    axis.text.y = element_text(size = 12),
  )

ggplotly(ranking_agreement_plot)

ggsave(" Agreement of Rankings.png", plot = ranking_scatter_plot, width = 10, height = 6, dpi = 300,bg="white")
```


## Alluvian plot 

```{r}
alluvial_data <- correctness_long %>%
  select(id, plot, phase, correctness) %>%
  pivot_wider(names_from = phase, values_from = correctness) %>%
  filter(!is.na(Pre_Learning) & !is.na(Post_Learning))  # Keep only complete pairs


# Alluvial diagram per plot
ggplot(alluvial_data,
       aes(axis1 = Pre_Learning, axis2 = Post_Learning, y = 1)) +
  geom_alluvium(aes(fill = Pre_Learning), width = 0.3) +
  geom_stratum(width = 0.3, fill = "gray80", color = "black") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2.3) +
  scale_x_discrete(limits = c("Pre_Learning", "Post_Learning"), expand = c(0.1, 0.1)) +
  facet_wrap(~plot) +
  labs(title = "Alluvial Diagram of Changes in Interpretation per Plot Type",
       x = "Phase",
       y = "Number of Respondents") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 8, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    strip.text = element_text(size = 12, face = "bold")
  )



```





#scatter plot with jitter and alpha blending (Ranking versus self ratede knowledge on survival )



# added groupng 

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)

# Prepare the data with proper ordering and grouping
ranking_scatter <- survey_data %>%
  select(id, knowledge_rate,
         ranking1_km, ranking2_km,
         ranking1_survdiff, ranking2_survdiff,
         ranking1_mrl, ranking2_mrl,
         ranking1_mr_ldiff, ranking2_mr_ldiff) %>%
  pivot_longer(
    cols = -c(id, knowledge_rate),
    names_to = c("phase", "plot"),
    names_pattern = "ranking(\\d)_(.*)",
    values_to = "rank"
  ) %>%
  mutate(
    phase = recode(phase, `1` = "Understand", `2` = "Consultation"),
    plot = case_when(
      plot == "km" ~ "Kaplan Meier",
      plot == "survdiff" ~ "Survival difference",
      plot == "mrl" ~ "MRL",
      plot == "mr_ldiff" ~ "MRL difference",
      TRUE ~ plot
    ),
    plot = factor(plot, levels = c("Kaplan Meier", "Survival difference", "MRL", "MRL difference")),
    rank = factor(rank, levels = c("1", "2", "3", "4")),
    knowledge_group = case_when(
      knowledge_rate < 5 ~ "Low",
      knowledge_rate >= 5 ~ "High"
    ),
    knowledge_group = factor(knowledge_group, levels = c("Low", "High"))
  ) %>%
  drop_na()

# Plot with grouped knowledge instead of raw score
ranking_grouped_plot <- ggplot(ranking_scatter, aes(x = knowledge_group, y = rank, color = phase, shape = phase)) +
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.6, size = 2.5) +
  
  facet_wrap(~plot, ncol = 2) +
  scale_x_discrete(name = "Self-rated Knowledge Group") +
  scale_y_discrete(name = "Rank (1 = Best)") +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Rank by Knowledge Group and Plot Type, Colored by Context") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_line(color = "gray85"),
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 13),
    axis.text = element_text(size = 11),
    strip.text = element_text(size = 13, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )



library(plotly)

ggplotly(ranking_grouped_plot)
```

 
 
 
## # ranking versus post learning score 
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(plotly)

# Replace 'post_score' with the actual column name representing post-learning score
ranking_scatter <- survey_data %>%
  select(id, post_score,
         ranking1_km, ranking2_km,
         ranking1_survdiff, ranking2_survdiff,
         ranking1_mrl, ranking2_mrl,
         ranking1_mr_ldiff, ranking2_mr_ldiff) %>%
  pivot_longer(
    cols = -c(id, post_score),
    names_to = c("phase", "plot"),
    names_pattern = "ranking(\\d)_(.*)",
    values_to = "rank"
  ) %>%
  mutate(
    phase = recode(phase, `1` = "Understand", `2` = "Consultation"),
    plot = case_when(
      plot == "km" ~ "Kaplan Meier",
      plot == "survdiff" ~ "Survival difference",
      plot == "mrl" ~ "MRL",
      plot == "mr_ldiff" ~ "MRL difference",
      TRUE ~ plot
    ),
    plot = factor(plot, levels = c("Kaplan Meier", "Survival difference", "MRL", "MRL difference")),
    rank = factor(rank, levels = c("1", "2", "3", "4")),
    score_group = case_when(
      post_score < 2 ~ "<2",
      post_score >= 2 ~ "≥2"
    ),
    score_group = factor(score_group, levels = c("<2", "≥2"),labels = c("below 50%","50% and Above"))
  ) %>%
  drop_na()

# Plot
ranking_grouped_plot <- ggplot(ranking_scatter, aes(x = score_group, y = rank, color = phase, shape = phase)) +
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.6, size = 2.5) +
  
  facet_wrap(~plot, ncol = 2) +
  scale_x_discrete(name = "Post-learning Score") +
  scale_y_discrete(name = "Rank (1 = Best)") +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Rank by Post-learning Score Group and Plot Type, Colored by Context") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_line(color = "gray85"),
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 13),
    axis.text = element_text(size = 11),
    strip.text = element_text(size = 13, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )

# Interactive version
ggplotly(ranking_grouped_plot)

```


## summary table 

```{r}

if (FALSE) {# Load required packages
library(tidyverse)
library(janitor)
library(readxl)
library(knitr)
library(kableExtra)
library(broom)

# Load data
survey_data <- read_xlsx("Survey_coded_export20250521.xlsx") %>%
  clean_names()

# Helper function: % accuracy
get_accuracy <- function(colname) {
  mean(survey_data[[colname]] == "CORRECT", na.rm = TRUE) * 100
}

# Helper function: Learning gain
get_learning_gain <- function(pre_col, post_col) {
  pre_acc <- get_accuracy(pre_col)
  post_acc <- get_accuracy(post_col)
  post_acc - pre_acc
}

# Helper function: McNemar p-value
get_learning_p <- function(pre_col, post_col) {
  tab <- table(survey_data[[pre_col]], survey_data[[post_col]])
  if (all(dim(tab) == c(2,2))) {
    pval <- broom::tidy(mcnemar.test(tab))$p.value
  } else {
    pval <- NA
  }
  return(pval)
}

# Helper function: % Top-ranked (Ranking == 1)
get_top_rank_pct <- function(rank_col) {
  mean(survey_data[[rank_col]] == 1, na.rm = TRUE) * 100
}

# Column mappings
plots <- c("KM Plot", "Survival Diff", "MRL Plot", "MRL Diff")
pre_cols <- c("pre_km_code", "pre_survdiff_code", "pre_mrl_code", "pre_mr_ldiff_code")
post_cols <- c("post_km_code", "post_survdiff_code", "post_mrl_code", "post_mr_ldiff_code")
rank1_cols <- c("ranking1_km", "ranking1_survdiff", "ranking1_mrl", "ranking1_mr_ldiff")
rank2_cols <- c("ranking2_km", "ranking2_survdiff", "ranking2_mrl", "ranking2_mr_ldiff")

# Build the table
summary_table <- tibble(
  Metric = c("Pre-learning accuracy", "Post-learning accuracy", "Learning gain (points)",
             "Learning gain p-value", "Top-ranked for patients", "Top-ranked for clinicians (ease of understanding)"),
  
  `KM Plot` = c(
    sprintf("%.1f%%", get_accuracy(pre_cols[1])),
    sprintf("%.1f%%", get_accuracy(post_cols[1])),
    sprintf("+%.1f", get_learning_gain(pre_cols[1], post_cols[1])),
    sprintf("p = %.3f", get_learning_p(pre_cols[1], post_cols[1])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[1])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[1]))
  ),
  
  `Survival Diff` = c(
    sprintf("%.1f%%", get_accuracy(pre_cols[2])),
    sprintf("%.1f%%", get_accuracy(post_cols[2])),
    sprintf("+%.1f", get_learning_gain(pre_cols[2], post_cols[2])),
    sprintf("p = %.3f", get_learning_p(pre_cols[2], post_cols[2])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[2])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[2]))
  ),
  
  `MRL Plot` = c(
    sprintf("%.1f%%", get_accuracy(pre_cols[3])),
    sprintf("%.1f%%", get_accuracy(post_cols[3])),
    sprintf("+%.1f", get_learning_gain(pre_cols[3], post_cols[3])),
    sprintf("p = %.3f", get_learning_p(pre_cols[3], post_cols[3])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[3])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[3]))
  ),
  
  `MRL Diff` = c(
    sprintf("%.1f%%", get_accuracy(pre_cols[4])),
    sprintf("%.1f%%", get_accuracy(post_cols[4])),
    sprintf("+%.1f", get_learning_gain(pre_cols[4], post_cols[4])),
    sprintf("p = %.3f", get_learning_p(pre_cols[4], post_cols[4])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[4])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[4]))
  )
)

# Render table
summary_table %>%
  kable("html", caption = "Summary of Key Outcomes", align = "lcccc") %>%
  kable_styling(full_width = FALSE, position = "center", bootstrap_options = c("striped", "hover"))}


```



```{r}
# Load required packages
library(tidyverse)
library(janitor)
library(readxl)
library(knitr)
library(kableExtra)
library(broom)
library(boot)

# Load data
survey_data <- read_xlsx("Survey_coded_export20250521.xlsx") %>%
  clean_names()

# Bootstrap helper functions

# Bootstrap for accuracy (mean proportion correct)
boot_mean <- function(data, indices) {
  sample_data <- data[indices]
  return(mean(sample_data == "CORRECT", na.rm = TRUE))
}

get_accuracy_boot <- function(colname, R = 5000) {
  vec <- survey_data[[colname]]
  vec <- vec[!is.na(vec)]
  
  boot_obj <- boot(data = vec, statistic = boot_mean, R = R)
  ci <- boot.ci(boot_obj, type = "perc")$percent[4:5]
  
  acc <- mean(vec == "CORRECT") * 100
  # Format: XX.X% [YY.Y - ZZ.Z]
  sprintf("%.1f%% [%.1f - %.1f]", acc, ci[1] * 100, ci[2] * 100)
}

# Bootstrap for learning gain
boot_learning_gain <- function(data, indices, pre_col, post_col) {
  sample_data <- data[indices, ]
  pre_acc <- mean(sample_data[[pre_col]] == "CORRECT", na.rm = TRUE)
  post_acc <- mean(sample_data[[post_col]] == "CORRECT", na.rm = TRUE)
  return(post_acc - pre_acc)
}

get_learning_gain_boot <- function(pre_col, post_col, R = 5000) {
  df <- survey_data %>% select(all_of(pre_col), all_of(post_col)) %>%
    drop_na()
  
  boot_obj <- boot(data = df, statistic = function(data, indices) {
    boot_learning_gain(data, indices, pre_col, post_col)
  }, R = R)
  
  ci <- boot.ci(boot_obj, type = "perc")$percent[4:5]
  
  gain <- mean(survey_data[[post_col]] == "CORRECT", na.rm = TRUE) * 100 -
          mean(survey_data[[pre_col]] == "CORRECT", na.rm = TRUE) * 100
  
  # Format: +XX.X [YY.Y - ZZ.Z]
  sprintf("+%.1f [%.1f - %.1f]", gain, ci[1] * 100, ci[2] * 100)
}

# McNemar p-value
get_learning_p <- function(pre_col, post_col) {
  tab <- table(survey_data[[pre_col]], survey_data[[post_col]])
  if (all(dim(tab) == c(2,2))) {
    pval <- broom::tidy(mcnemar.test(tab))$p.value
  } else {
    pval <- NA
  }
  return(pval)
}

# % Top-ranked (Ranking == 1)
get_top_rank_pct <- function(rank_col) {
  mean(survey_data[[rank_col]] == 1, na.rm = TRUE) * 100
}

# Column mappings
plots <- c("KM Plot", "Survival Diff", "MRL Plot", "MRL Diff")
pre_cols <- c("pre_km_code", "pre_survdiff_code", "pre_mrl_code", "pre_mr_ldiff_code")
post_cols <- c("post_km_code", "post_survdiff_code", "post_mrl_code", "post_mr_ldiff_code")
rank1_cols <- c("ranking1_km", "ranking1_survdiff", "ranking1_mrl", "ranking1_mr_ldiff")
rank2_cols <- c("ranking2_km", "ranking2_survdiff", "ranking2_mrl", "ranking2_mr_ldiff")

# Build improved summary table
summary_table <- tibble(
  Metric = c("Pre-learning accuracy", "Post-learning accuracy", "Learning gain (points)",
             "Learning gain p-value", "Top-ranked for patients", "Top-ranked for clinicians (ease of understanding)"),
  
  `KM Plot` = c(
    get_accuracy_boot(pre_cols[1]),
    get_accuracy_boot(post_cols[1]),
    get_learning_gain_boot(pre_cols[1], post_cols[1]),
    sprintf("p = %.3f", get_learning_p(pre_cols[1], post_cols[1])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[1])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[1]))
  ),
  
  `Survival Diff` = c(
    get_accuracy_boot(pre_cols[2]),
    get_accuracy_boot(post_cols[2]),
    get_learning_gain_boot(pre_cols[2], post_cols[2]),
    sprintf("p = %.3f", get_learning_p(pre_cols[2], post_cols[2])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[2])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[2]))
  ),
  
  `MRL Plot` = c(
    get_accuracy_boot(pre_cols[3]),
    get_accuracy_boot(post_cols[3]),
    get_learning_gain_boot(pre_cols[3], post_cols[3]),
    sprintf("p = %.3f", get_learning_p(pre_cols[3], post_cols[3])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[3])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[3]))
  ),
  
  `MRL Diff` = c(
    get_accuracy_boot(pre_cols[4]),
    get_accuracy_boot(post_cols[4]),
    get_learning_gain_boot(pre_cols[4], post_cols[4]),
    sprintf("p = %.3f", get_learning_p(pre_cols[4], post_cols[4])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[4])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[4]))
  )
)

# Render table
summary_table %>%
  kable("html", caption = "Summary of Key Outcomes (with 95% Bootstrap CIs)", align = "lcccc") %>%
  kable_styling(full_width = FALSE, position = "center", bootstrap_options = c("striped", "hover"))


if (FALSE) {escape_percent <- function(x) {
  stringr::str_replace_all(x, "%", "\\\\%")
}

summary_table_safe <- summary_table %>%
  mutate(across(-Metric, escape_percent))




summary_table_safe %>%
  kable("latex", booktabs = TRUE, escape = TRUE, caption = "Summary of Key Outcomes (with 95\\% Bootstrap CIs)", align = "lcccc") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))}



```

```{r}
# Helper function to escape % signs for LaTeX
escape_percent <- function(x) {
  stringr::str_replace_all(x, "%", "\\\\%")
}

# Build improved summary table (same as before)
summary_table <- tibble(
  Metric = c("Pre-learning accuracy", "Post-learning accuracy", "Learning gain (points)",
             "Learning gain p-value", "Top-ranked for patients", "Top-ranked for clinicians (ease of understanding)"),
  
  `KM Plot` = c(
    get_accuracy_boot(pre_cols[1]),
    get_accuracy_boot(post_cols[1]),
    get_learning_gain_boot(pre_cols[1], post_cols[1]),
    sprintf("p = %.3f", get_learning_p(pre_cols[1], post_cols[1])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[1])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[1]))
  ),
  
  `Survival Diff` = c(
    get_accuracy_boot(pre_cols[2]),
    get_accuracy_boot(post_cols[2]),
    get_learning_gain_boot(pre_cols[2], post_cols[2]),
    sprintf("p = %.3f", get_learning_p(pre_cols[2], post_cols[2])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[2])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[2]))
  ),
  
  `MRL Plot` = c(
    get_accuracy_boot(pre_cols[3]),
    get_accuracy_boot(post_cols[3]),
    get_learning_gain_boot(pre_cols[3], post_cols[3]),
    sprintf("p = %.3f", get_learning_p(pre_cols[3], post_cols[3])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[3])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[3]))
  ),
  
  `MRL Diff` = c(
    get_accuracy_boot(pre_cols[4]),
    get_accuracy_boot(post_cols[4]),
    get_learning_gain_boot(pre_cols[4], post_cols[4]),
    sprintf("p = %.3f", get_learning_p(pre_cols[4], post_cols[4])),
    sprintf("%.0f%%", get_top_rank_pct(rank2_cols[4])),
    sprintf("%.0f%%", get_top_rank_pct(rank1_cols[4]))
  )
)

# Now apply escape_percent() to all columns except "Metric"
summary_table_safe <- summary_table %>%
  mutate(across(-Metric, escape_percent)) %>% 
  mutate(across(-Metric, as.character))

# Render table (Elsevier-friendly LaTeX version)
summary_table_safe %>%
  kable("latex", booktabs = TRUE, escape = TRUE,
        caption = "Summary of Key Outcomes (with 95\\% Bootstrap CIs)",
        align = "lcccc") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
```




## Ranking chnages (allevial plot)
```{r}

library(ggalluvial)


# Create long format for rankings
ranking_long <- survey_data %>%
  mutate(id = row_number()) %>%
  pivot_longer(
    cols = starts_with("ranking"),
    names_to = "ranking_type",
    values_to = "rank"
  ) %>%
  mutate(
    plot_type = case_when(
      str_detect(ranking_type, "km") ~ "Kaplan Meier(KM)",
      str_detect(ranking_type, "survdiff") ~ "Survival difference",
      str_detect(ranking_type, "mrl$") ~ "Mean Residual Life(MRL)",
      str_detect(ranking_type, "mr_ldiff") ~ "MRL difference"
    ),
    preference = case_when(
      str_detect(ranking_type, "ranking1") ~ "Pre",
      str_detect(ranking_type, "ranking2") ~ "Post"
    )
  ) %>%
  drop_na(rank)

# Reshape into wide format to define flows: one row per person-plot
ranking_wide <- ranking_long %>%
  pivot_wider(id_cols = c(id, plot_type),
              names_from = preference,
              values_from = rank)

# Plot: Alluvial diagram
ranking_change_plot <- ggplot(ranking_wide,
       aes(axis1 = as.factor(Pre),
           axis2 = as.factor(Post))) +
  geom_alluvium(aes(fill = plot_type), width = 0.25, alpha = 0.8) +
  geom_stratum(width = 0.25, color = "black", fill = "gray90") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  facet_wrap(~plot_type) +
  scale_x_discrete(limits = c("Easy to understand", "use with patient")) +
  labs(
    title = "Ranking Changes: Easy to understand by clinician versus Use with patient in consultation",
    x = "preference", y = "Number of Participants"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(face = "bold", size = 14))

ggsave("Ranking change.png", plot = ranking_change_plot, width = 10, height = 6, dpi = 300,bg="white")
```



## Fitting a tree to the data set #####################################################


```{r}

library(rpart.plot)
# Create useful predictors
tree_data <- survey_data %>%
  mutate(
    pre_total_correct = rowSums(across(c(pre_km_code, pre_survdiff_code, pre_mrl_code, pre_mr_ldiff_code), ~ . == "CORRECT")),
    post_mrl_correct = factor(post_mrl_code, levels = c("INCORRECT", "CORRECT")),
    qualification = factor(qualification),
    Self_rated_knowledge = as.numeric(knowledge_rate)
  ) %>%
  select(post_mrl_correct, Self_rated_knowledge, pre_total_correct, qualification) %>%
  drop_na()

tree_data <- tree_data %>%
  mutate(
    qualification = case_when(
      qualification == "Medical Doctor (Specialist or postgraduate)" ~ "MD_Specialist",
      qualification == "Medical Doctor (General Practitioner)" ~ "MD_GP",
      qualification == "Medical student" ~ "Med_student"
    )
  )
head(tree_data)
# Fit decision tree model
tree_model <- rpart(post_mrl_correct ~ ., data = tree_data, method = "class", control = rpart.control(cp = 0, minsplit = 10,minbucket =5 ,maxdepth = 10))

# Plot the tree
rpart.plot(tree_model, type = 4, extra = 104,
           main = "Prediction of correct MRL plot interpretation post learning section")

```


## predict learning gain 

```{r}


# data process
tree_data_gain <- survey_data %>%
  mutate(
    # Convert correctness codes to numeric (1 = correct, 0 = incorrect)
    pre_total_correct = rowSums(across(c(pre_km_code, pre_survdiff_code, pre_mrl_code, pre_mr_ldiff_code), ~ . == "CORRECT")),
    post_total_correct = rowSums(across(c(post_km_code, post_survdiff_code, post_mrl_code, post_mr_ldiff_code), ~ . == "CORRECT")),
    learning_gain = post_total_correct - pre_total_correct,
    qualification = case_when(
      qualification == "Medical Doctor (Specialist or postgraduate)" ~ "MD_Specialist",
      qualification == "Medical Doctor (General Practitioner)" ~ "MD_GP",
      qualification == "Medical student" ~ "Med_student"
    ),
    knowledge_rate = as.numeric(knowledge_rate)
  ) %>%
  select(learning_gain, knowledge_rate, pre_total_correct, qualification) %>%
  drop_na()

# Fit regression tree model
reg_tree <- rpart(learning_gain ~ ., data = tree_data_gain, method = "anova", control = rpart.control(cp = 0, minsplit = 10,minbucket =5 ,maxdepth = 10))

# Plot the regression tree
 rpart.plot(reg_tree, type = 4, extra = 101,
           main = "Regression Tree Predicting Learning Gain")


```
######################################################
```{r}
write.csv(survey_data, "survey_data_shiny.csv",row.names = FALSE)
```
#############################################
