---
title: "Victor_Thesis"
author: "Victor Rwandarwacu"
date: "2024-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TCGA Clinical data set

https://portal.gdc.cancer.gov/projects/TCGA-BRCA 

```{r }

library(survival)
library(survminer)
library(dplyr)
library(ggplot2)
library(tidyverse)
```

## Dataset overview 

This is a data set from NIH National Cancer Institute , TCGA Program on a project called "Breast invasive carcinoma (BRCA)"
The data set contains information about: demography, exposure , Family History(regarding cancer), Follow up, Molecular Test, other Clinical Attribute, pathology detail,and Treatment.

```{r , echo=FALSE}
# Import the TSV file TCGA brca clinical data 

brca_clinical <- read.delim("clinical.tsv")
#sapply(brca_clinical,class)

names(brca_clinical)
head(brca_clinical)
nrow(brca_clinical)
```


##subgroug needed for survival analysis ####

This survival analysis included the following columns: gender, race, ethnicity  age at diagnosis, vital status (Survival state), cancer staging and Management.

```{r , echo=FALSE}
brca_cli_sub<- brca_clinical|>select(case_id,gender,race,ethnicity,age_at_index, year_of_birth,year_of_death,age_at_diagnosis,vital_status,days_to_diagnosis,days_to_death,days_to_last_follow_up,ajcc_pathologic_stage, ajcc_pathologic_t, ajcc_pathologic_n,ajcc_pathologic_m,treatment_type,)

head(brca_cli_sub)
names(brca_clinical)

```
##data cleaning 

Record variables names  and types 

```{r , echo=FALSE}


 
##rename column 
colnames(brca_cli_sub)<- c("case_id","gender","race","ethnicity","age_at_index", "year_of_birth","year_of_death","age_at_diagnosis","vital_status","days_to_diagnosis","days_to_death","days_to_last_follow_up", "pathologic_stage","pathologic_t","pathologic_n","pathologic_m","treatment_type")


```







```{r , echo=FALSE}
brca_cli_sub<- brca_cli_sub|> mutate(age_at_diagnosis=as.numeric(age_at_diagnosis),days_to_death=as.numeric(days_to_death),days_to_last_follow_up=as.numeric(days_to_last_follow_up))

#conversion of time from days to years 
#brca_cli_sub<- brca_cli_sub|>mutate(age_diag_yrs=age_at_diagnosis/365.25,yrs_to_event=days_to_death/365.25,yrs_to_last_follow_up=days_to_last_follow_up/365.25)


```



```{r , echo=FALSE}
###remain with high level stages 

brca_cli_sub <- brca_cli_sub %>%
  mutate(pathologic_stage = case_when(
    grepl("Stage ([IVX]+)[A-Z]*", pathologic_stage) ~ gsub("Stage ([IVX]+)[A-Z]*", "Stage \\1", pathologic_stage),
    TRUE ~ "Unspecified"
  ))



head(brca_cli_sub)
```



```{r , echo=FALSE}
#modification of the treatment type into 2 categories

brca_cli_sub <- brca_cli_sub %>%
  mutate(treatment_type = case_when(
    grepl("Pharmaceutical Therapy, NOS", treatment_type, ignore.case = TRUE) ~ "Chemotherapy",
    grepl("Radiation Therapy, NOS",treatment_type, ignore.case = TRUE) ~ "Radiotherapy",
    TRUE ~ "Other"))
```







```{r , echo=FALSE}
#  Modify "race" to include only three categories
brca_cli_sub <- brca_cli_sub %>%
  mutate(race = case_when(
    grepl("black or african american", race, ignore.case = TRUE) ~ "Black or African American",
    grepl("white", race, ignore.case = TRUE) ~ "White",
    TRUE ~ "Other"
  ))

##remove male patients 
#brca_cli_sub <- brca_cli_sub %>%filter(tolower(gender) != "male"| is.na(gender))
##remove negative time to follow up
#brca_cli_sub <- brca_cli_sub %>%filter(days_to_last_follow_up >= 0)
#brca_cli_sub <- brca_cli_sub %>%filter(days_to_last_follow_up >= 0,yrs_to_last_follow_up>=0)
summary(brca_cli_sub)
```



```{r , echo=FALSE}


```


##distribution of variables
distribution of age at diagnosis 

```{r , echo=FALSE}
brca_cli_sub %>%
  ggplot(aes(x = age_at_diagnosis)) +
  geom_histogram(binwidth = 2, na.rm = TRUE, fill = "blue", color = "black") +
  labs(
    title = "Distribution of Age at Diagnosis",
    x = "Age at Diagnosis (Years)",
    y = "Count"
  )



brca_cli_sub %>%
  ggplot(aes(x = days_to_death)) +
  geom_histogram(bins = 30, na.rm = TRUE, fill = "blue", color = "black") +
  labs(
    title = "Distribution of time to event",
    x = "time to event (Years)",
    y = "Count"
  )

brca_cli_sub %>%
  ggplot(aes(x = days_to_last_follow_up)) +
  geom_histogram(bins = 30, na.rm = TRUE, fill = "blue", color = "black") +
  labs(
    title = "Distribution of time of follow up",
    x = "time of follow up (Years)",
    y = "Count"
  )
summary(brca_cli_sub$age_diag_yrs)
```







##contigency tables##

```{r , echo=FALSE}
addmargins(table(brca_cli_sub$vital_status,brca_cli_sub$treatment_type))

```



```{r , echo=FALSE}
addmargins(table(brca_cli_sub$race,brca_cli_sub$vital_status))
```
there is predominance of the white population.

encode the vital status column  and add plot KMP AND cumulative functions plot 


###merging the time to event and time to follow up data##########################################################


```{r , echo=FALSE}
#USING IF ELSE 
#brca_cli_sub<- brca_cli_sub|> mutate(time= if_else(is.na(days_to_death),days_to_last_follow_up,days_to_death))

#USING RESHAPE
#transformation to long format 
brca_cli_sub <- brca_cli_sub %>%
  mutate(id = row_number())

brca_cli_sub_long<- reshape(
  brca_cli_sub,
  varying = list(c("days_to_last_follow_up", "days_to_death")),
  v.names = "time",
  timevar = "source",
  times = c("days_to_last_follow_up", "days_to_death"),
  direction = "long"
)
head(brca_cli_sub_long)
#filling the missing values in long format 

filled_time<- brca_cli_sub_long %>%
  group_by(id) %>%
  summarize(time = first(na.omit(time)), .groups = "drop")

#reshape back to wide format , 
#data set with merged time 

brca_cli_sub_filled<-brca_cli_sub|> select(-days_to_death,-days_to_last_follow_up) |> left_join(filled_time, by = "id")
head(brca_cli_sub_filled)
```




```{r , echo=FALSE}

##keep only female patients 
brca_cli_sub_filled <- brca_cli_sub_filled %>%filter(tolower(gender) == "female")
##Keep needed variables only 
##remove negative time 
brca_cli_sub_filled <- brca_cli_sub_filled %>%filter(time >= 0)

#sort the data set by treatment, to keep individual patient instead of duplicates  , for further analysis, 
brca_cli_sub_chemo <- brca_cli_sub_filled|> filter(treatment_type == "Chemotherapy")



```



```{r }
####Survival data set 

surv_brca_cli<- brca_cli_sub_chemo|> select(gender,race,ethnicity,vital_status,pathologic_stage,pathologic_t,pathologic_n,pathologic_m,time) 

surv_brca_cli <- surv_brca_cli |> mutate(time_in_years = time /365.25 )

summary(surv_brca_cli)
hist(surv_brca_cli$time_in_years)

```





###preprocessing of survival data#### 
```{r , echo=FALSE}
#change columns from characters to factors 

surv_brca_cli[] <- lapply(surv_brca_cli, function(x) if(!is.numeric(x)) as.factor(x) else x)
sapply(surv_brca_cli,class)

#encode vital status 
surv_brca_cli <- surv_brca_cli|> mutate( vital_status= ifelse(vital_status == "Alive", 0, 1))

colSums(is.na(surv_brca_cli))
write.csv(surv_brca_cli, file = "surv_brca_cli.csv", row=FALSE)

summary(surv_brca_cli)
```


##Survival object##

Overall Mortality risk and implications

The following table include mortality  summary and risk 
The graph is the kaplan maier plot of all values 
```{r , echo=FALSE}
#survival object for use 

surv_object <- Surv(time = surv_brca_cli$time_in_years, event = surv_brca_cli$vital_status)


fit1 <- survfit(surv_object ~ 1, data = surv_brca_cli)
mortality_summary <- data.frame(
    Time = summary(fit1)$time,
    Risk = 1 - summary(fit1)$surv,
    Std.Error = summary(fit1)$std.err,
    Lower.95CI = summary(fit1)$lower,
    Upper.95CI = summary(fit1)$upper
  )

head(mortality_summary)


ggsurvplot(
  fit1,
  conf.int = TRUE,                
  risk.table = TRUE,              
  xlab = "Time in years",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),
  
)  
```


##KM PLOT stratification by pathologic stages 


```{r , echo=FALSE}
#fit the kaplan maier model 
fit_path_stage <- survfit(surv_object ~ pathologic_stage , data = surv_brca_cli)
#plot the survival curve 
summary(fit_path_stage)

ggsurvplot(
  fit_path_stage,
  size = 1.5,                  # Line thickness
  font.x = c(16, "bold"),      # Font size for X-axis
  font.y = c(16, "bold"),      # Font size for Y-axis
  font.tickslab = c(10),       # Font size for tick labels
  legend = "right",             # Legend position
  conf.int = TRUE,                
  risk.table = TRUE,             
  pval = TRUE,                    
  xlab = "Time(years)",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),      
  legend.title = "pathological stage",     
  risk.table.height = 0.5,        
  surv.median.line = "hv" 
  
)

```


```{r , echo=FALSE}
survdiff(formula = surv_object ~ pathologic_stage,  data = surv_brca_cli)


```

## survival ratio
survival ratio plot for the pathological stage 
This a novel approach of visualization of the survival data compring different covariates.
For demonstration we will compare stage II and III

```{r , echo=FALSE}
#data frame with pathological stages as co-variate 
##extraction of stage 3 and 2 data frame from the data set 


path_stageII_df<- surv_brca_cli|> select(time,vital_status,pathologic_stage)|> filter(pathologic_stage == "Stage II")
path_stageIII_df<- surv_brca_cli|> select(time,vital_status,pathologic_stage)|> filter(pathologic_stage == "Stage III")

#KM fit of the data 
#surv_ob_Path_stageII<- Surv(time = surv_ratio_brca_cli$time,event = surv_ratio_brca_cli$vital_status)

km_fit_path_stageII <- survfit(Surv(time= time, event=vital_status) ~ 1, data = path_stageII_df)
km_fit_path_stageIII <- survfit(Surv(time= time, event=vital_status) ~ 1, data = path_stageIII_df)

####extract survival probabilities 

# Manually define time points (e.g., every 30 days)
manual_times <- seq(0, max(path_stageII_df$time, na.rm = TRUE), by = 30)

# Extract survival probabilities for Pathologic stage 2
surv_pathII <- summary(km_fit_path_stageII, times = manual_times, extend = TRUE)

# Extract survival probabilities for pathologic stage 3
surv_pathIII <- summary(km_fit_path_stageIII, times = manual_times, extend = TRUE)


# Survival ratios and confidence intervals
survival_ratio <- surv_pathIII$surv / surv_pathII$surv


lower_ratio <- surv_pathIII$lower / surv_pathII$upper
upper_ratio <- surv_pathIII$upper / surv_pathII$lower

# Combine into a data frame
ratio_data <- data.frame(
  Time = manual_times,
  Path_stageIII_surv= surv_pathIII$surv,
  Path_stageII_surv= surv_pathII$surv,
  Survival_Ratio = survival_ratio,
  Lower = lower_ratio,
  Upper = upper_ratio
)

# Survival probabilities for stage 2 and 3  (for plotting individual curves)
surv_pathII_curve <- data.frame(Time = surv_pathII$time, Survival = surv_pathII$surv, Group = "pathologic_stageII")
surv_pathIII_curve <- data.frame(Time = surv_pathIII$time, Survival = surv_pathIII$surv, Group = "pathologic_stageIII")

# Combine survival curves into a single data frame
survival_curves_path23_combined <- rbind(surv_pathII_curve, surv_pathIII_curve)

library(ggplot2)

# Plot survival ratio and individual survival curves
ggplot() +
  # Plot individual survival curves
  geom_step(data = survival_curves_path23_combined, aes(x = Time, y = Survival, color = Group), size = 1) +
  # Plot survival ratio
  geom_step(data = ratio_data, aes(x = Time, y = Survival_Ratio), color = "green", size = 1) +
  geom_ribbon(data = ratio_data, aes(x = Time, ymin = Lower, ymax = Upper), alpha = 0.3, fill = "red") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "blue") +  # Reference line at ratio = 1
  labs(
    title = "Survival Ratio and Kaplan-Meier Curves",
    x = "Follow-Up Time (Days)",
    y = "Survival Probability / Ratio",
    color = "Group"
  ) +
  annotate("text", x = max(ratio_data$Time) * 0.3, y = 2, label = "Higher survival for stage 3", hjust = 0) +
  annotate("text", x = max(ratio_data$Time) * 0.3, y = 0.5, label = "Higher survival for stage 2", hjust = 0) +
  theme_minimal() +
  scale_color_manual(values = c("pathologic_stageII" = "blue", "pathologic_stageIII" = "red"))

```
```{r}
#check significance for the plots stage 2 and stage 3  
path_stage2_3_df<- surv_brca_cli|> select(time,vital_status,pathologic_stage)|> filter(pathologic_stage == "Stage II" | pathologic_stage == "Stage III")
head(path_stage2_3_df)
fit23 <- survfit(Surv(time,vital_status)~pathologic_stage, data = path_stage2_3_df)
LogRank23<- survdiff(Surv(time,vital_status)~pathologic_stage, data = path_stage2_3_df)
summary(fit23)
summary(LogRank23)

ggsurvplot(
  fit23,
  conf.int = TRUE,                
  risk.table = TRUE,              
  xlab = "Time in days",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),
  pval = TRUE
) 
```


##Methods other than sequence 
```{r}

stageII_surv <- summary(
  fit_path_stage,
  times = fit_path_stage$time[fit_path_stage$strata == "pathologic_stage = StageII"],
 
)

stageII_times <- fit_path_stage$time[fit_path_stage$strata == "pathologic_stage=Stage II"]
print(stageII_times)



summary(fit_path_stage)
fit_path_stage$time[fit_path_stage$strata=="pathologic_stage=Stage II"]

```




##parametric modelling for cancer stages#
generate an new data set for survreg

```{r , echo=FALSE}
surv_brca_cli$pathologic_stage <- as.factor(surv_brca_cli$pathologic_stage)
summary(surv_brca_cli$time)
regsurv_brca_cli <- surv_brca_cli[surv_brca_cli$time > 0, ]

reg_path_stage <- survreg(Surv(surv_brca_cli$time,surv_brca_cli$vital_status) ~ pathologic_stage , data = regsurv_brca_cli, dist="weibull")
#plot the survival curve 
summary(reg_path_stage)

table(surv_brca_cli$vital_status)
table(surv_brca_cli$pathologic_stage)


#######################################################################################################



# Generate survival probabilities from the parametric model
new_data <- data.frame(pathologic_stage = levels(regsurv_brca_cli$pathologic_stage)) # Modify for covariates
times <- seq(0, max(regsurv_brca_cli$time), by = 1)  # Define time range
###option 1 
surv_curves <- sapply(1:nrow(new_data), function(i) {
  predict(reg_path_stage, newdata = new_data[i, ,drop=FALSE ], type = "quantile", p = seq(0, 1, length.out = length(times)))
})

# Plot the curves using ggplot2
library(ggplot2)
df <- data.frame(
  Time = rep(times, each = nrow(new_data)),
  Survival = c(t(surv_curves)),
  Group = rep(new_data$pathologic_stage, each = length(times))
)

ggplot(df, aes(x = Time, y = Survival, color = Group)) +
  geom_line(linewidth = 1.5) +
  labs(
    title = "Parametric Survival Curves",
    x = "Time (years)",
    y = "Survival Probability",
    color = "Pathological Stage"
  ) +
  theme_minimal()
#option 2 

surv_curves2 <- sapply(1:nrow(new_data), function(i) {
  sapply(times, function(t) {
    predict(
      reg_path_stage,
      newdata = new_data[i, , drop = FALSE],
      type = "response",
      p = t
    )
  })
})

df <- data.frame(
  Time = rep(times, nrow(new_data)),
  Survival = c(t(surv_curves)),  # Flatten the matrix into a vector
  Group = rep(new_data$pathologic_stage, each = length(times))
)

ggplot(df, aes(x = Time, y = Survival, color = Group)) +
  geom_line(size = 1.5) +
  labs(
    title = "Parametric Survival Curves",
    x = "Time (years)",
    y = "Survival Probability",
    color = "Pathological Stage"
  ) +
  theme_minimal()








```














#stratification of survival  by ethnicity 

```{r , echo=FALSE}

#fit the kaplan meier model 


#fit the kaplan maier model 
fit_ethn <- survfit(surv_object ~ ethnicity , data = surv_brca_cli)
#plot the survival curve 

ggsurvplot(
  fit_ethn,
  size = 1.5,                  # Line thickness
  font.x = c(10, "bold"),      # Font size for X-axis
  font.y = c(10, "bold"),      # Font size for Y-axis
  font.tickslab = c(10),       # Font size for tick labels
  legend = "right",             # Legend position
  conf.int = TRUE,                
  risk.table = TRUE,             
  pval = TRUE,                    
  xlab = "Time(years)",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),      
  legend.title = "race",     
  risk.table.height = 0.3,        
  surv.median.line = "hv" 
  
)







```






#survival stratification by Race 

```{r , echo=FALSE}

#fit the kaplan maier model 
fit_race <- survfit(surv_object ~ race , data = surv_brca_cli)
#plot the survival curve 

ggsurvplot(
  fit_race,
  size = 1.5,                  
  font.x = c(12, "bold"),     
  font.y = c(12, "bold"),     
  font.tickslab = c(10),       
  legend = "right",             
  conf.int = TRUE,                
  risk.table = TRUE,             
  pval = TRUE,                    
  xlab = "Time(years)",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),      
  legend.title = "race",     
  risk.table.height = 0.3,        
  surv.median.line = "hv" 
  
)

```


median survival time black is 17.5 years 


```{r , echo=FALSE}
survdiff(formula = surv_object ~ race,  data = surv_brca_cli)


```
##stratification by metastasis 

```{r , echo=FALSE}

#fit the kaplan maier model 
fit_path_n <- survfit(surv_object ~ pathologic_n , data = surv_brca_cli)
#plot the survival curve 

ggsurvplot(
  fit_path_n,
  size = 1.5,                  
  font.x = c(12, "bold"),     
  font.y = c(12, "bold"),     
  font.tickslab = c(10),       
  legend = "right",             
  conf.int = TRUE,                
  risk.table = TRUE,             
  pval = TRUE,                    
  xlab = "Time(years)",                  
  ylab = "Survival Probability",  
  ggtheme = theme_minimal(),      
  legend.title = "Path. ",     
  risk.table.height = 0.3,        
  surv.median.line = "hv" 
  
)





survdiff(formula = surv_object ~ pathologic_m,  data = surv_brca_cli)

```







##log rank test 

Log rank test is used  to compare the distribution of time to event data between 2 indipendent samples.
 $H_{0}$ is that both curves has identical distribution (all curves are the same), 
 $H_{A}$ curves has different distribution 
 



```{r , echo=FALSE}



```

